---
title: "ECMWF Data - Preliminary Look"
output: html_document
date: "2023-09-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

- The aim of this document is a quick exploration of ECMWF Seasonal Forecast data. The data was downloaded with  `data-raw/ecmwf_seas51.R` using the `{ecmwfr}` package.

- Currently awaiting inputs from partners regarding the preferred data sources. Therefore, this document is a preliminary look at the data without a very clear objective rather than familiarization. 

- There is an ongoing larger discussion regarding converting monthly forecast values to data to trimester tercile probability values. This is out of the scope of this document.

**NOTE:** IMO - no need for code review at te moment

```{r cars}
library(tidyverse)
library(tidync)
library(rhdx)
library(sf)
library(rnaturalearth)
library(ncdf4)
library(terra)
library(exactextractr)
library(targets)
library(glue)
gghdx()

tar_load(gdf_aoi_adm)


fp_seas51 <- file.path(
  Sys.getenv("AA_DATA_DIR"),
  "private",
  "raw",
  "lac",
  "ecmwf_seasonal",
  "seas51"
)

# lets simplify admin 1 - remove slivers/islands
adm0_main <- gdf_aoi_adm$adm0 %>%
  st_cast("POLYGON") %>%
  mutate(
    area = as.numeric(st_area(.))
  ) %>%
  filter(
    area > 1e10
  )

poly_simp_adm0 <- adm0_main %>%
  group_by(adm0_es) %>%
  summarise() %>%
  st_simplify(dTolerance = 0.01)



fnames <- dir(fp_seas51)

# list of rasters
lr <- fnames %>%
  map(
    \(fname){
      fp <- file.path(fp_seas51, fname)
      nf <- nc_open(fp)

      # get precipitation rate
      # it's 4D for each value there is 1. lat, 2. lon, 3. ensemble, 4. time)
      tprate_array <- ncvar_get(nf, "tprate")

      # get fill value
      fill_value <- ncatt_get(nf, "tprate", "_FillValue")

      # fill array w/ standard NA
      tprate_array[tprate_array == fill_value$value] <- NA
      tprate_mean_array <- apply(tprate_array, c(1, 2, 4), mean, na.rm = T)


      lon <- ncvar_get(nf, "longitude")
      lat <- ncvar_get(nf, "latitude", verbose = F)
      t_hours <- ncvar_get(nf, "time")

      r <- rast(
        x = aperm(tprate_mean_array, c(2, 1, 3)),
        extent = ext(
          min(lon) - .5,
          max(lon) + .5,
          min(lat) - .5,
          max(lat) + 0.5
        ),
        crs = "EPSG:4326"
      )
      dates_predicted <- as_date(as_date("1900-01-01") + hours(t_hours))
      set.names(r, dates_predicted)
      return(r)
    }
  ) %>%
  set_names(fnames)
```


```{r}
# run zonal stats at regional level (all 4 countries)

# list rasters historical
lr_historical <- lr %>%
  keep_at(~ str_detect(.x, "2022"))

# list rasters latest
lr_latest <- lr %>%
  keep_at(~ str_detect(.x, "2023"))



df_zonal_historical_regional <- map2_dfr(
  lr_historical,
  c(lt = 1, lt = 2, lt = 3, lt = 4),
  \(rtmp, lt){
    exact_extract(
      x = rtmp,
      y = poly_simp_adm0 %>%
        summarise(),
      fun = "mean"
    ) %>%
      pivot_longer(everything()) %>%
      separate(name, into = c("stat", "date"), sep = "\\.") %>%
      pivot_wider(names_from = "stat", values_from = "value") %>%
      mutate(
        date = as_date(date),
        leadtime = lt
      )
  }
)

df_zonal_latest_regional <- map2_dfr(
  lr_latest,
  c(lt = 1, lt = 2, lt = 3, lt = 4),
  \(rtmp, lt){
    exact_extract(
      x = rtmp,
      y = poly_simp_adm0 %>%
        summarise(),
      fun = "mean"
    ) %>%
      pivot_longer(everything()) %>%
      separate(name, into = c("stat", "date"), sep = "\\.") %>%
      pivot_wider(names_from = "stat", values_from = "value") %>%
      mutate(
        date = as_date(date),
        leadtime = lt
      )
  }
)

df_zonal_regional <- bind_rows(
  df_zonal_historical_regional,
  df_zonal_latest_regional,
) %>%
  mutate(
    # units are m/s - convert to meters: 60 s * 60 minutes * 24 hours * 30 days
    # meters to mm = m * 1000
    total_mean_mm = (mean * (60 * 60 * 24 * 30)) * 1000
  )
```


```{r}
df_zonal_historical_adm0 <- map2_dfr(
  lr_historical,
  c(lt = 1, lt = 2, lt = 3, lt = 4),
  \(rtmp, lt){
    exact_extract(
      x = rtmp,
      y = poly_simp_adm0,
      fun = "mean",
      append_cols = "adm0_es"
    ) %>%
      pivot_longer(-matches("adm0_es")) %>%
      separate(name, into = c("stat", "date"), sep = "\\.") %>%
      pivot_wider(names_from = "stat", values_from = "value") %>%
      mutate(
        date = as_date(date),
        leadtime = lt
      )
  }
)

df_zonal_latest_adm0 <- map2_dfr(
  lr_latest,
  c(lt = 1, lt = 2, lt = 3, lt = 4),
  \(rtmp, lt){
    exact_extract(
      x = rtmp,
      y = poly_simp_adm0,
      fun = "mean",
      append_cols = "adm0_es"
    ) %>%
      pivot_longer(-matches("adm0_es")) %>%
      separate(name, into = c("stat", "date"), sep = "\\.") %>%
      pivot_wider(names_from = "stat", values_from = "value") %>%
      mutate(
        date = as_date(date),
        leadtime = lt
      )
  }
)

df_zonal_adm0 <- bind_rows(
  df_zonal_historical_adm0,
  df_zonal_latest_adm0
) %>%
  mutate(
    # units are m/s - convert to meters: 60 s * 60 minutes * 24 hours * 30 days
    # meters to mm = m * 1000
    total_mean_mm = (mean * (60 * 60 * 24 * 30)) * 1000
  )
```

# ECMWF Monthly Forecast values

To simplify visual we just look at leadtime 1 values for all historical forecast records. Values summarised to 
```{r}

df_zonal_adm0 %>%
  filter(leadtime == 1) %>%
  ggplot(
    aes(
      x = date,
      y = total_mean_mm,
      group = adm0_es
    )
  ) +
  geom_point(size= 1,alpha= 0.5) +
  geom_line(alpha=0.5) +
  facet_wrap(~adm0_es)+
  labs(
    title = "Mean Historical Precipitation Forecast (ECMWF)",
    subtitle="Only leadtime 1 shown for simplification of visual",
    y= "precipitation (mm)",
    caption = "Source: ECMWF Monthly Forecast"
  )+
  theme(
    axis.title.x = element_blank()
  )
  
```

Not a very useful visual - but leaving here:

```{r}
df_zonal_adm0 %>%
  ggplot(
    aes(
      x = date,
      y = total_mean_mm,
      group = adm0_es
    )
  ) +
  geom_point(size= 1,alpha= 0.5) +
  geom_line(alpha=0.5) +
  facet_wrap(~adm0_es)+
  labs(
    title = "Mean Historical Precipitation Forecast (ECMWF)",
    subtitle="Across all leadtimes",
    y= "precipitation (mm)",
    caption = "Source: ECMWF Monthly Forecast"
  )+
  facet_grid(
    # vars(adm0_es, leadtime),
    cols=vars(leadtime),
    rows=vars(adm0_es)
    )+
  theme(
    axis.title.x = element_blank()
  )
```

# Dryest Years as predicted by ECMWF
```{r}
df_zonal_regional_seas <- df_zonal_regional %>%
  mutate(
    mo = month(date),
    yr = year(date)
  ) %>%
  mutate(
    season = case_when(
      mo %in% c(5, 6, 7) ~ "MJJ",
      mo %in% c(6, 7, 8) ~ "JJA",
      mo %in% c(9, 10, 11) ~ "SON",
      .default = "no_window"
    ),
    season = fct_relevel(season, "MJJ", "JJA", "SON"),
    yr_season = paste0(yr, "_", season)
  ) %>%
  # just do lt 1 for reference -- probably should do across all.
  filter(leadtime == 1)

df_min_per_season <- df_zonal_regional_seas %>%
  group_by(yr_season, yr, season) %>%
  summarise(
    min = min(total_mean_mm),
    avg = mean(total_mean_mm), .groups = "drop"
  )


df_min_of_min <- df_min_per_season %>%
  group_by(season) %>%
  slice_min(order_by = min, n = 5) %>%
  ungroup() %>%
  filter(season != "no_window")

df_mean_of_min <- df_min_per_season %>%
  group_by(season) %>%
  slice_min(order_by = avg, n = 5) %>%
  ungroup() %>%
  filter(season != "no_window")



df_el_nino_years <- tibble(
  date =
    c(
      "1982-1983",
      "1997-1998",
      "2002-2003",
      "2004-2005",
      "2006-2007",
      "2009-2010",
      "2014-2016",
      "2018-2019",
      "2023-2024"
    )
) %>%
  separate(
    date,
    into = c("start_date", "end_date"), sep = "-"
  )

df_el_nino_years <- df_el_nino_years %>%
  mutate(
    start_date = glue("{start_date}-01-01") %>% as_date(),
    end_date = glue("{end_date}-12-31") %>% as_date()
  )



df_min_per_season %>%
  filter(season != "no_window") %>%
  mutate(
    mo = case_when(
      str_detect(season, "JJA") ~ "07",
      str_detect(season, "MJJ") ~ "03",
      str_detect(season, "SON") ~ "09",
    ),
    date = as_date(glue("{yr}-{mo}-01"))
  ) %>%
  mutate(
    pt_label = ifelse(yr_season %in% df_mean_of_min$yr_season, yr, NA) %>%
      as_factor(),
    pt_color = yr_season %in% df_mean_of_min$yr_season
  ) %>%
  ggplot() +
  scale_x_date(
    date_breaks = "2 year",
    date_labels = "%Y"
  ) +
  scale_y_reverse() +
  geom_point(
    aes(
      x = date,
      y = avg,
      color = pt_color
    )
  ) +
  geom_text(
    aes(
      label = pt_label,
      x = date,
      y = avg,
      color = pt_color
    ),
    color = hdx_hex("tomato-hdx"), nudge_y = 20
  ) +
  geom_rect(
    data = df_el_nino_years,
    aes(
      xmin = start_date,
      xmax = end_date,
      ymin = -Inf,
      ymax = Inf
    ),
    alpha = 0.2,
    fill = hdx_hex("tomato-hdx")
  ) +
  facet_wrap(~season, nrow = 33) +
  labs(
    title = "Dryest seasons according to ECMWF (LT 1) forecasts",
    subtitle = "Central American Dry Corridor (el Nino years highlighted in red)",
    y = "Monthly Average Precipitation (mm)"
  ) +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(), 
    panel.border = element_rect(color = "grey", fill = NA),
    axis.text.x = element_text(angle = 90)
  )
```
