---
title: "IRI_VHI_Trigger_Proposal_Revised V1"
format: html
editor: visual
---

## Intro

This document will attempt to propose a v1 trigger for all 3 monitoring windows using just IRI seasonal forecast and VHI (FAO) data. In v2 of the proposal we may want to also consider ECMWF seasonal foreast data.

From previous analysis it's fairly clear that we can aggregate VHI (by median value) for each country and set a reasonable threshold between 40 & 50. Due to lack of historical IRI data this threshold is not well informed.

FAO has recommended also including VHI over cropland as an indicator and has recommended a threshold of ≥25 % of cropland with a VHI < 0.35. This is based on some global analysis they have peformed. However, all the details are not yet available. 

Therefore let's more critically examine the VHI indicator.


## VHI

Come up w/ 3 year RP joint prob. However, if using IRI data we don;t have long enough record to justify lowering it due to VHI, Therefore let's go with an extremem 5-10 year RP for VHI that will trigger it on it's own or allow for a just over bavg forecastr -- have to be pick in months

## IRI approx 42.5

Boom


## IRI 40 + VHI 5 year


```{r}
library(targets)
library(janitor)
library(tidyverse)
tar_load(df_vhi_median_over_crop)
df_chirps <- read_csv(file.path(
  Sys.getenv("AA_DATA_DIR"),
  "public",
  "processed",
  "lac",
  "chirps_monthly_adm0.csv"
))



```


## Driest years
```{r}
df_vhi_chirps <- bind_rows(
  df_vhi_median_over_crop %>% 
    mutate(
      date= as_date(paste0(date,"-01")),
      indicator = "Median VHI"
    ),
  df_chirps %>% 
    clean_names() %>% 
    select(adm0_es=adm0_name, stat=parameter, date ,value) %>% 
    mutate(
      indicator = "monthly precip"
    )
)



df_vhi_chirps %>% 
  filter(indicator =="monthly precip") %>% 
  ggplot(
    aes(
      x= date, y= value
    )
    
  )+
  geom_line(color =hdx_hex("sapphire-dark"),
    alpha=0.5)+
  geom_line(
    data= df_vhi_chirps %>% 
  filter(indicator =="Median VHI"),
  aes(x= date, y= value *500), color=hdx_hex("mint-dark"),alpha=0.8
  )+
  scale_y_continuous(name = "monthly precip",
                     sec.axis = sec_axis(~./500, name = "Median VHI")
                     )+
  facet_wrap(~adm0_es)


df_chirps_yearly_historical <- df_vhi_chirps %>% 
  filter(
    indicator =="monthly precip",
    year(date)<2023
         ) %>% 
  group_by(yr= floor_date(date,"year"),adm0_es) %>% 
  
  summarise(
    mm_yearly = sum(value),.groups="drop"
  )

driest_yrs<- df_chirps_yearly_historical %>% 
  group_by(adm0_es) %>% 
  slice_min(n = 5,order_by = mm_yearly)

p_yearly<- df_chirps_yearly_historical %>% 
  ggplot(
    aes(x= yr,
        y= mm_yearly, 
        color = adm0_es,
        group=adm0_es)
  )+
  geom_line()+
  geom_point(
    data= driest_yrs,
    aes(x= yr, y= mm_yearly)
  )

p_yearly

```

 Table
```{r}
gt_table_driest_years <-  df_chirps_yearly_historical %>% 
  group_by(adm0_es) %>% 
   arrange(
     mm_yearly
   ) %>% 
   mutate(
     rank = row_number()
   ) %>% 
  pivot_wider(
    names_from = adm0_es,
    values_from = mm_yearly:rank
  ) %>% 
   arrange(yr) %>% 
  gt() %>% 
  gt::cols_hide(
    starts_with("rank")
    ) %>% 
   gt::cols_label(
     yr="Year",
    mm_yearly_Honduras= "Honduras",
    `mm_yearly_El Salvador`= "El Salvador",
    `mm_yearly_Nicaragua`= "Nicaragua",
    `mm_yearly_Guatemala`= "Guatemala",
    
   ) %>% 
  tab_header(
    title = "Driest years by country"
  ) %>% 
  data_color(
        columns= c("mm_yearly_El Salvador"),
        rows = `rank_El Salvador`<=5,
        method = "numeric",
        palette = "YlOrRd",
        reverse=T
      ) %>% 
  data_color(
        columns= c("mm_yearly_Guatemala"),
        rows = `rank_Guatemala`<=5,
        method = "numeric",
        palette = "YlOrRd",
        reverse=T
      ) %>% 
  data_color(
        columns= c("mm_yearly_Honduras"),
        rows = `rank_Honduras`<=5,
        method = "numeric",
        palette = "YlOrRd",
        reverse=T
      ) %>% 
  data_color(
        columns= c("mm_yearly_Nicaragua"),
        rows = `rank_Nicaragua`<=5,
        method = "numeric",
        palette = "YlOrRd",
        reverse=T
      ) 


gt_table_driest_years
  
```

```{r}
df_chirps_yearly_historical %>% 
  ggplot(
    aes(x= mm_yearly)
  )+
  geom_histogram()+
  facet_wrap(~adm0_es)

yearly_lower_33<- df_chirps_yearly_historical %>% 
  group_by(adm0_es) %>% 
  summarise(
    q33= quantile(mm_yearly,0.33),
  )

df_chirps_classified<- df_chirps %>% 
  clean_names() %>% 
  group_by(adm0_es=adm0_name, yr= floor_date(date,"year")) %>% 
  mutate(
   mm_yearly = sum(value)
 ) %>% 
  ungroup() %>% 
  left_join(yearly_lower_33) %>% 
  mutate(
    drier_third =mm_yearly<= q33
  )

df_chirps_classified %>% 
  group_by(
    adm0_es,drier_third, month = month(date, label =T)
  ) %>% 
  summarise(
    mm = mean(value)
  ) %>% 
  ggplot(
    aes(x= month, y= mm)
  )+
  geom_bar(
    aes(fill= drier_third), stat= "identity",position="dodge"
  )

df_chirps_classified %>% 
  group_by(
    adm0_es, mo = month(date, label =T)
  ) %>% 
  slice_min(
    order_by = value,n=3
  ) %>% 
  select(adm0_es, mo, date,value) %>% 
  filter(
    mo %in% c("Jul","Jun","Aug")
  ) %>% 
  arrange(mo) %>% 
  print(n=nrow(.))

df_chirps_windows <- df_chirps_classified %>% 
  group_by(
    lgl_mjj =month(date)%in% c(5,6,7),
    lgl_jja =month(date)%in% c(6,7,8),
    lgl_son =month(date)%in% c(9,10,11),
  )

df_chirps_windows_summed <- c("lgl_mjj","lgl_jja","lgl_son") %>% 
  map(
    ~df_chirps_windows %>% 
      group_by(
        across(starts_with("adm")),
        yr_date=floor_date(date,"year"),
        !!sym(.x)
      ) %>% 
      summarise(
        mm = sum(value),.groups="drop"
      ) %>% 
      filter(!!sym(.x)) %>% 
      mutate(
        window = .x
      ) %>% 
      select(-any_of(.x))
  ) %>% 
  list_rbind()



lower_third_windows <- df_chirps_windows_summed %>% 
  group_by(adm0_es, window) %>% 
  summarise(
    q_33= quantile(mm,0.33)
  )

df_chirps_windows_summed_classified <- df_chirps_windows_summed %>% 
  left_join(
    lower_third_windows
  ) %>% 
  mutate(
    lgl_q33 = mm<=q_33
  ) 

df_chirps_windows_summed_classified %>% 
  ggplot(
    aes(x= yr_date, y= mm)
  )+
  geom_line()+
  geom_point(aes(col= lgl_q33))+
  facet_grid (
    cols = vars(window),
    rows= vars(adm0_es)
  )


df_chrirps_windows_wide <- df_chirps_windows_summed_classified %>% 
  ungroup() %>% 
  select(-adm0_name) %>% 
  group_by(adm0_es, window) %>% 
  arrange(mm) %>% 
  mutate(
    rank = row_number()
  ) %>% 
  select(
    adm0_es, yr_date,window, rank,mm,
  ) %>% 
  pivot_wider(
    names_from = c("window","adm0_es"),
    values_from=c("rank","mm")
  ) %>% 
  arrange(yr_date) 
  df_chrirps_windows_wide %>% 
    glimpse()
  
mute_palette <-  "Oranges"
mute_alpha <- 0.1

  
df_chrirps_windows_wide %>% 
  gt() %>% 
  #   tab_style(
  #   style = list(
  #     cell_text(color = "black", weight = "bold")
  #   )
  # )
  cols_hide(columns= starts_with("rank")) %>% 
  cols_label(
    yr_date="Year",
    mm_lgl_mjj_Honduras= "MJJ (window 1)",
    mm_lgl_jja_Honduras= "JJA (window 2)",
    mm_lgl_son_Honduras= "SON (window 3)",
    mm_lgl_mjj_Nicaragua= "MJJ (window 1)",
    mm_lgl_jja_Nicaragua= "JJA (window 2)",
    mm_lgl_son_Nicaragua= "SON (window 3)",
    mm_lgl_mjj_Guatemala= "MJJ (window 1)",
    mm_lgl_jja_Guatemala= "JJA (window 2)",
    mm_lgl_son_Guatemala= "SON (window 3)",
    `mm_lgl_mjj_El Salvador`= "MJJ (window 1)",
    `mm_lgl_jja_El Salvador`= "JJA (window 2)",
    `mm_lgl_son_El Salvador`= "SON (window 3)",
  ) %>%
  tab_spanner(columns = c("mm_lgl_mjj_Honduras","mm_lgl_jja_Honduras","mm_lgl_son_Honduras"), label = "Honduras") %>% 
  tab_spanner(columns = c("mm_lgl_mjj_Nicaragua","mm_lgl_jja_Nicaragua","mm_lgl_son_Nicaragua"), label = "Nicaragua") %>% 
  tab_spanner(columns = c("mm_lgl_mjj_Guatemala","mm_lgl_jja_Guatemala","mm_lgl_son_Guatemala"), label = "Guatemala") %>% 
  tab_spanner(columns = c("mm_lgl_mjj_El Salvador","mm_lgl_jja_El Salvador","mm_lgl_son_El Salvador"), label = "El Salvador") %>% 
  # tab_spanner(columns = ends_with("Nicaragua"), label = "Nicarauga") %>% 
  # tab_spanner(columns = ends_with("Guatemala"), label = "Guatemala") %>% 
  # tab_spanner(columns = ends_with("El Salvador"), label = "El Salvador") %>% 
  tab_header(
    title = "Driest years by country"
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_Honduras"),
    rows = `rank_lgl_mjj_Honduras`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_Honduras"),
    rows = `rank_lgl_mjj_Honduras`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_Honduras"),
    rows = `rank_lgl_jja_Honduras`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_Honduras"),
    rows = `rank_lgl_jja_Honduras`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_Honduras"),
    rows = `rank_lgl_son_Honduras`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_Honduras"),
    rows = `rank_lgl_son_Honduras`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_Nicaragua"),
    rows = `rank_lgl_mjj_Nicaragua`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_Nicaragua"),
    rows = `rank_lgl_mjj_Nicaragua`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_Nicaragua"),
    rows = `rank_lgl_jja_Nicaragua`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_Nicaragua"),
    rows = `rank_lgl_jja_Nicaragua`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_Nicaragua"),
    rows = `rank_lgl_son_Nicaragua`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_Nicaragua"),
    rows = `rank_lgl_son_Nicaragua`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_Guatemala"),
    rows = `rank_lgl_mjj_Guatemala`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_Guatemala"),
    rows = `rank_lgl_mjj_Guatemala`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_Guatemala"),
    rows = `rank_lgl_jja_Guatemala`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_Guatemala"),
    rows = `rank_lgl_jja_Guatemala`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_Guatemala"),
    rows = `rank_lgl_son_Guatemala`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_Guatemala"),
    rows = `rank_lgl_son_Guatemala`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_El Salvador"),
    rows = `rank_lgl_mjj_El Salvador`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_mjj_El Salvador"),
    rows = `rank_lgl_mjj_El Salvador`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_El Salvador"),
    rows = `rank_lgl_jja_El Salvador`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_jja_El Salvador"),
    rows = `rank_lgl_jja_El Salvador`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_El Salvador"),
    rows = `rank_lgl_son_El Salvador`<=5,
    method = "numeric",
    palette = "YlOrRd",
    alpha=1,
    reverse=T
  ) %>% 
  data_color(
    columns= c("mm_lgl_son_El Salvador"),
    rows = `rank_lgl_son_El Salvador`>5,
    method = "numeric",
    palette = mute_palette,
    alpha=mute_alpha,
    reverse=T
  ) 

  glimpse()

  split(.$adm0_es) %>% 
  map(
    \(dft){
      dft %>% 
        select(-lgl_q33, -q_33) %>% 
        pivot_wider(
          names_from = window,
          values_from=mm
        ) %>% 
        gt() %>% 
        cols_label(
          yr_date="Year",
          lgl_mjj= "MJJ (window 1)",
          lgl_jja= "JJA (window 2)",
          lgl_son= "SON (window 3)",
        ) %>%
        cols_hide(columns =c("adm0_code","adm0_name")) %>% 
        tab_header(
          title = "Driest years by country"
        ) %>% 
        data_color(
          columns= c("lgl_mjj","lgl_jja","lgl_son"),
          # rows = `rank_El Salvador`<=5,
          method = "numeric",
          palette = "YlOrRd",
          alpha=0.5,
          reverse=T
        ) 
    }
  )

gt_chirps_windows$Nicaragua
gt_chirps_windows$Guatemala
gt_chirps_windows$Honduras
gt_chirps_windows$`El Salvador`

```

```{r}

select(
    date,
    starts_with("adm"),
    starts_with("lgl"),
    value
  ) %>% 
  ungroup() %>% 
  filter(
    if_any(starts_with("lgl"),~.x==T)
  ) %>% 
  pivot_longer(
    cols = starts_with("lgl"),
    # names_to = "window",
    values_to = "value"
  ) 
    
  )
  pivot_longer(
    cols = starts_with("lgl"),
    names_to = "window",
    values_to = "lgl"
  ) %>% 
  group_by(
    adm0_es,yr, window
  ) %>% 
  summarise(
    mm= sum(value)
  ) 
  



summarise(
    mm = mean(value)
  ) %>% 
  ggplot(
    aes(x= month, y= mm)
  )+
  geom_bar(
    aes(fill= drier_third), stat= "identity",position="dodge"
  )+
  facet_wrap(~adm0_es)

  
  
```


```{r}
df_vhi_chirps_anom <- df_vhi_chirps %>% 
  mutate(
    mo = month(date,label=T)
  ) %>% 
  group_by(
    adm0_es,  indicator,mo
  ) %>% 
  mutate(
    mean =mean(value),
    stdev = sd(value),
    diff_mean = value - mean,
    z_score = diff_mean/stdev
    
  )

df_vhi_chirps_anom %>% 
  ggplot(
    aes(
      x= date, y= z_score, color =indicator, group=indicator
    )
    
  )+
  geom_line(
    alpha=0.5)+
  scale_y_continuous(
                     
                     )+
  facet_wrap(~adm0_es)

drought_years <-  c(2014, 2018)

df_vhi_chirps_anom %>% 
  select(adm0_es, indicator, date, z_score) %>% 
  mutate(
    in_drought = year(date)%in% drought_years
  ) %>% 
  pivot_wider(
    names_from = indicator,
    values_from = z_score
  ) %>% 
  ggplot(
    aes(x= `monthly precip`, y= `Median VHI`, color=in_drought)
  )+
  geom_point()+
  facet_wrap(
    ~adm0_es
  )

```

Since VHI is observed we can only look-back so what are the months we would be looking back at?

- First trigger window is MJJ
  + VHI not included in monitoring framework
- Second trigger window is JJA
  + at end of May/early July we can look back at **May** & **April**
- Third trigger window is SON
  + at end of August/early September we can look back at **August** & **July**

```{r}
# window 2
  
df_vhi_relevant_windows <- df_vhi_chirps_anom %>% 
  mutate(
    window = case_when(
      mo %in% c("May","Apr") ~ "window 2",
      mo %in% c("Aug","Jul") ~ "window 3"
    )
  ) %>% 
  filter(!is.na(window),
         indicator =="Median VHI")





df_vhi_freq_thresh <- 
  map2(
    \(thresh_temp){
      df_vhi_relevant_windows %>% 
        split(,$window) %>% 
        pivot_wider
       mutate(
         lgl_lte = value <= thresh_temp
       ) %>% 
       group_by(adm0_es,window,mo) %>%
        summarise(
          freq_pct = sum(lgl_lte)/n(),
          freq_count = sum(lgl_lte),.groups="drop"
        ) %>% 
        mutate(
          thresh= thresh_temp,
        )
    }
  ) %>% 
  list_rbind()

target_rp <-  3
target_pct <-  1/target_rp




# filter to values closest to target pct
df_thresh_optimal <- df_vhi_freq_thresh %>% 
  mutate(
    diff = abs(freq_pct - target_pct)
  ) %>% 
  arrange(diff) %>% 
  group_by(window,adm0_es) %>% 
  slice_min(n = 2, order_by = diff) %>% 
  slice_min(n= 1, order_by= thresh) %>% 
  mutate(
    thresh_uid = paste0(adm0_es, window, mo)
  )
  
  

df_vhi_relevant_windows %>% 
  ungroup() %>% 
  mutate(
    thresh_uid = paste0(adm0_es, window, mo),
    lgl_breach = value < filter(df_thresh_optimal, thresh_uid == thresh_uid)$thresh
  ) %>% 
  ggplot(
    aes(
      x= date, y= value
    )
  )+
  geom_line()+ 
  geom_point(
    aes(color=lgl_breach)
  )+
  scale_color_manual(values=c(hdx_hex("mint-hdx"),hdx_hex("tomato-hdx")))+
  facet_grid(
    cols = vars(adm0_es),
    rows = vars(window)
  )


```

## Joint 
```{r}

df_vhi_mean_sd <- df_vhi_relevant_windows %>% 
  group_by(
    across(starts_with("adm")),
    window, 
    mo = month(date)
  ) %>% 
  summarise(
    mean = mean(value),
    stdev = sd(value)
  )
df_vhi_window_z_min <- df_vhi_relevant_windows %>% 
  group_by(
    across(starts_with("adm")),
    window, 
    yr= year(date)
  ) %>% 
  summarise(
    z_score =min(z_score)
  )
z_range <- df_vhi_window_z_min %>% 
  pull(z_score) %>% 
  range()

df_vhi_relevant_windows %>% 
  ggplot(
    aes(x= z_score)
  )+
  geom_histogram()+
  facet_grid(rows= vars(adm0_es),cols=vars(window))




thresholds_quantiile <- df_vhi_window_z_min %>% 
   group_by(
    across(starts_with("adm")),
    window) %>% 
  summarise(
    q_z_33 = quantile(z_score,0.33)
    
  ) %>% 
  left_join(
    df_vhi_mean_sd %>% 
      group_by(
        across(starts_with("adm")),
        window
      ) %>% 
      arrange(
        mo
      ) %>% 
      mutate(
        mo_order = row_number()
      ) %>% 
      select(-mo) %>% 
      pivot_wider(
        names_from = mo_order,
        values_from = c(mean,stdev),
        names_glue = "{.value}_{mo_order}"
      )
    
  ) %>% 
  mutate(
    q_abs_33_month1 = (q_z_33*stdev_1)+mean_1,
    q_abs_33_month2 = (q_z_33*stdev_2)+mean_2
  )
  


df_vhi_relevant_windows %>% 
  left_join(thresholds_quantiile)



seq(z_range[1], z_range[2], by =0.05) %>% 
  map(\(thresh_temp){
    df_vhi_relevant_windows %>% 
                                     mutate(
                                       z_score = z<= t1,
                                       month_2_lte = month_2<= t2,
                                       month_1_or_2_lte = month_1_lte|month_2_lte
                                     ) 
  }
    
  )
```


```{r}

thresh_combos <- expand_grid(
  thresh_1 = seq(0,1,0.01),
  thresh_2 = seq(0,1,0.01)
)

df_vhi_relevant_for_thresholding <- df_vhi_relevant_windows %>% 
  ungroup() %>% 
  mutate(
    yr = year(date)
  ) %>% 
  select(starts_with("adm"),yr,value,mo,window) %>% 
  split(.$window) %>% 
  map(
    \(dft){
      dft %>% 
        pivot_wider(
    names_from = "mo",
    values_from = "value",
  ) %>% 
        
        rename(
          month_1 = 5,
          month_2 = 6
        )
    }
  ) %>% 
  list_rbind() 

df_vhi_conditional_thresh<- map2(thresh_combos$thresh_1, 
                                 thresh_combos$thresh_2,
                                 \(t1,t2){
                                   df_vhi_relevant_for_thresholding %>% 
                                     mutate(
                                       month_1_lte = month_1<= t1,
                                       month_2_lte = month_2<= t2,
                                       month_1_or_2_lte = month_1_lte|month_2_lte
                                     ) %>% 
                                     group_by(adm0_es,window) %>%
                                     summarise(
                                       conditonal_pct = sum(month_1_or_2_lte)/n(),
                                       conditional_count = sum(month_1_or_2_lte),
                                       mo1_pct = sum(month_1_lte)/n(),
                                       mo1_count = sum(month_1_lte),
                                       mo2_pct = sum(month_2_lte)/n(),
                                       mo2_count = sum(month_2_lte),
                                       .groups="drop"
                                     ) %>% 
                                     mutate(
                                       thresh1 = t1,
                                       thresh2 = t2
                                     )
                                 }) %>% 
  list_rbind()

df_vhi_conditional_thresh %>% 
  count(conditonal_pct) %>% 
  print(n=nrow(.))
df_vhi_conditional_thresh %>% 
  mutate(
    diff_conditional_targ = abs(conditonal_pct - target_pct)
  ) %>% 
  group_by(adm0_es,window) %>% 
  slice_min(order_by = diff_conditional_targ,with_ties = T
  ) %>% 
  View()


# filter to values closest to target pct
df_conditional_thresh_optimal <- df_vhi_conditional_thresh %>% 
  mutate(
    diff = abs(freq_pct - target_pct),
    thresh_diff= abs(thresh1 - thresh2)
  ) %>% 
  group_by(window,adm0_es) %>% 
  filter(between(freq_pct,0.32,0.34)) %>% 
  arrange(diff, thresh_diff)

df_conditional_thresh_optimal %>% 
  print(n=500)

df_conditional_thresh_optimal %>% 
  filter(window=="window 2") %>% 
  filter(between(freq_pct,0.3,0.4)) %>% 
  ggplot(
    aes(x= thresh1, y= thresh2)
  )+
  geom_point()+
  scale_x_continuous(breaks = seq(0,0.5,by =0.1))+
  scale_y_continuous(breaks = seq(0,0.5,by =0.1))+
  facet_wrap(
    ~adm0_es
  )
  

```



```{r}



