---
title: "Explore INSIVUMEH Special Forecast 2026"
format: html
---

## Objective

Identify which PRCP.nc file to use as the forecast by listing available files and extracting model codes.

```{r}
library(tidyverse)

# List all files in the special forecast directory
forecast_dir <- file.path(
  Sys.getenv("AA_DATA_DIR_NEW"),
  "private/raw/lac/INSIVUMEH/special_forecast_2026"
)

files <- list.files(forecast_dir, full.names = FALSE)

# Store in data.frame and extract model code from PRCP files
df_files <- tibble(
  filename = files
) |>
  mutate(
    is_prcp = str_detect(filename, "_PRCP\\.nc$"),
    model_code = if_else(
      is_prcp,
      str_extract(filename, "[A-Za-z0-9]+_PRCP\\.nc$"),
      NA_character_
    )
  )

df_files
```

```{r}
# Filter to PRCP files, extract year, and count per year/model
df_prcp <- df_files |>
  filter(is_prcp) |>
  mutate(
    year = str_extract(filename, "\\d{4}") |> as.integer()
  )

df_prcp |>
  count(year, model_code) |>
  pivot_wider(names_from = model_code, values_from = n, values_fill = 0)
```

```{r}

# look - Marzo is the correct spelling in the other variables, There are just no PRCP.nc files for Marzo
df_files |>
  mutate(
    year = str_extract(filename, "\\d{4}") |> as.integer(),
    month = str_extract(filename, "^[A-Za-z]+")
  ) |> count(month)

df_files |>
  filter(str_detect(tolower(filename),"mar")) |> 
  print(n=156)
  

```

```{r}
# Extract month from filename and check which are missing
all_months <- c("Enero", "Febrero", "Marzo", "Abril", "Mayo", "Junio",
                "Julio", "Agosto", "Septiembre", "Octubre", "Noviembre", "Diciembre")

df_prcp <- df_prcp |>
  mutate(
    month = str_extract(filename, "^[A-Za-z]+")
  )

# Find missing months per year/model
df_missing <- df_prcp |>
  group_by(year, model_code) |>
  summarise(
    missing_months = list(setdiff(all_months, month)),
    n_missing = length(setdiff(all_months, month)),
    .groups = "drop"
  ) |>
  filter(n_missing > 0) |>
  unnest(missing_months)

# Create complete grid to show presence/absence
month_abbr <- c("E", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
names(month_abbr) <- all_months

df_complete <- expand_grid(
  year = unique(df_prcp$year),
  model_code = unique(df_prcp$model_code),
  month = all_months
) |>
  left_join(
    df_prcp |> select(year, model_code, month) |> mutate(present = TRUE),
    by = c("year", "model_code", "month")
  ) |>

  mutate(
    present = replace_na(present, FALSE),
    month = factor(month, levels = all_months)
  )

# Heatmap - compact visualization for email
df_complete |>
  mutate(model_code = str_remove(model_code, "_PRCP\\.nc$")) |>
ggplot(aes(x = month, y = factor(year), fill = present)) +

geom_tile(color = "white", linewidth = 0.5) +
  facet_wrap(~model_code, ncol = 2) +
  scale_fill_manual(
    values = c("TRUE" = "#2E7D32", "FALSE" = "#C62828"),
    labels = c("Faltante", "Disponible")
  ) +
  scale_x_discrete(labels = month_abbr) +
  labs(
    title = "Disponibilidad de Archivos PRCP de INSIVUMEH por Modelo",
    x = "Mes", y = "Año", fill = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 9),
    legend.position = "bottom"
  )
```

```{r}
# Compact summary table: count of available months per year/model
df_complete |>
  group_by(year, model_code) |>
  summarise(n_available = sum(present), .groups = "drop") |>
  pivot_wider(names_from = model_code, values_from = n_available) |>
  arrange(year)
```

## Check leadtimes

Each file should contain 6 leadtimes. Let's verify by extracting publication date (S) and valid date (T) from all PRCP files.

```{r}
library(ncmeta)
library(lubridate)

# Extract pub/valid dates and calculate leadtime for all PRCP files
df_leadtimes <- df_prcp |>
  mutate(fp = file.path(forecast_dir, filename)) |>
  rowwise() |>
  mutate(
    t_units = nc_att(fp, "T", "units")$value[[1]],
    s_units = nc_att(fp, "S", "units")$value[[1]],
    valid_date = as.Date(str_extract(t_units, "\\d{4}-\\d{2}-\\d{2}")),
    pub_date = as.Date(str_extract(s_units, "\\d{4}-\\d{2}-\\d{2}")),
    lt = interval(pub_date, valid_date) %/% months(1)
  ) |>
  ungroup() |>
  select(filename, model_code, pub_date, valid_date, lt)

# Leadtime distribution
df_leadtimes |>
  count(lt, name = "n_files")

df_leadtimes |> 
  count(month(pub_date))
```

```{r}
# Summary
cat("Total PRCP files:", nrow(df_leadtimes), "\n")
cat("Unique leadtimes:", paste(unique(df_leadtimes$lt), collapse = ", "), "\n")
```

**Finding:** All files contain only leadtime = 1 (1-month ahead forecasts). Filename indicates the target/valid month (e.g., "Abril2000" = forecast FOR April, published in March). No files with leadtimes 2-6 are present.

## Vizual for explaining issue


```{r}
# For a given target month, we expect 6 files (LT 1-6)
# Example: For Mayo 2020, we expect:
#   - LT1: published April 2020
#   - LT2: published March 2020
#   - LT3: published February 2020
#   - LT4: published January 2020
#   - LT5: published December 2019
#   - LT6: published November 2019

# Show what we have vs what we expect for a sample year
sample_year <- 2020
expected_leadtimes <- 1:6

sample_model <- "CFSv2"

df_lt_check <- df_leadtimes |>
  filter(year(valid_date) == sample_year,
         str_detect(model_code, sample_model)) |>
  mutate(valid_month = month(valid_date, label = TRUE, abbr = FALSE)) |>
  select(valid_month, lt) |>
  distinct()

# Create expected grid
df_expected <- expand_grid(

  valid_month = month(1:12, label = TRUE, abbr = FALSE),
  lt_expected = expected_leadtimes
) |>
  left_join(
    df_lt_check |> mutate(present = TRUE),
    by = c("valid_month", "lt_expected" = "lt")
  ) |>
  mutate(present = replace_na(present, FALSE))

# Heatmap: valid month vs leadtime
ggplot(df_expected, aes(x = factor(lt_expected), y = valid_month, fill = present)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = if_else(present, "\u2714", "\u2718")), size = 5, family = "Arial Unicode MS") +
  scale_fill_manual(
    values = c("TRUE" = "#2E7D32", "FALSE" = "#FFCDD2"),
    labels = c("Faltante", "Disponible")
  ) +
  labs(
    title = paste0("Leadtimes disponibles por mes válido (", sample_year, ", modelo ", sample_model, ")"),
    subtitle = "Solo tenemos leadtime = 1. Faltan leadtimes 2-6 para todos los meses.",
    x = "Leadtime (meses de anticipación)",
    y = "Mes válido (target)",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
```

```{r}
# Simple summary table to send to provider
df_leadtimes |>
  mutate(
    valid_month = format(valid_date, "%B %Y"),
    pub_month = format(pub_date, "%B %Y")
  ) |>
  count(lt, name = "n_archivos") |>
  mutate(
    explicacion = case_when(
      lt == 1 ~ "Pronóstico publicado 1 mes antes del mes válido",
      lt == 2 ~ "Pronóstico publicado 2 meses antes del mes válido",
      TRUE ~ paste0("Pronóstico publicado ", lt, " meses antes")
    )
  ) |>
  knitr::kable(
    col.names = c("Leadtime", "N° Archivos", "Explicación"),
    caption = "Resumen: Todos los 827 archivos PRCP tienen leadtime = 1"
  )
```

```{r}
# Concrete example for one target month and one model
ejemplo <- df_leadtimes |>
  filter(month(valid_date) == 5, year(valid_date) == 2020,
         str_detect(model_code, "CFSv2")) # Mayo 2020, CFSv2

cat("Ejemplo concreto: Pronósticos para Mayo 2020 (modelo CFSv2)\n")
cat("=" |> rep(55) |> paste(collapse = ""), "\n\n")

cat("Lo que TENEMOS:\n")
ejemplo |>
  mutate(info = paste0("  - ", filename, " (pub: ", pub_date, ", LT=", lt, ")")) |>
  pull(info) |>
  cat(sep = "\n")

cat("\n\nLo que ESPERAMOS (6 leadtimes):\n")
cat("  - LT1: publicado Abril 2020     [SI] (disponible)\n")
cat("  - LT2: publicado Marzo 2020     [NO] (faltante)\n")
cat("  - LT3: publicado Febrero 2020   [NO] (faltante)\n")
cat("  - LT4: publicado Enero 2020     [NO] (faltante)\n")
cat("  - LT5: publicado Diciembre 2019 [NO] (faltante)\n")
cat("  - LT6: publicado Noviembre 2019 [NO] (faltante)\n")
```
