---
title: "VHI"
output: html_document
date: "2023-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

Visualize VHI as % of cropland per country.

We will first think of this in the context of the framework w/ the 2 thresholds provided by partners:

- VHI threshold= 0.35
- % Area ≤ VHI threshold (0.35) = 25%

Then we will evaluate other threshold combinations.

**Takeaway** After evaluating all threshold combinations, those provided by partners seem reasonable:

```{r cars}
library(targets)
library(tidyverse)
library(gghdx)
gghdx()
tar_load(df_cropland_lte_vhi_threshold)

vhi_thresh <- 0.35
area_thresh <- 0.25


df_vhi_area_thresh <- df_cropland_lte_vhi_threshold %>%
  ungroup() %>%
  filter(round(threshold, 2) == vhi_thresh) %>%
  mutate(
    date = ym(date),
    flag_area_thresh = pct_crop_pixels_lte_thresh >= area_thresh
  )
```


## Using Proposed Thresholds

# Flagging Months

```{r}
df_vhi_area_thresh %>%
  ggplot(
    aes(x = date, y = pct_crop_pixels_lte_thresh)
  ) +
  geom_point(aes(color = flag_area_thresh)) +
  # geom_line()+
  geom_hline(yintercept = area_thresh, linetype = "dashed", color = "tomato") +
  scale_y_continuous_hdx(labels = scales::label_percent()) +
  labs(
    y = "% Cropland",
    title = "Vegetative Health Index - Proposed Threshold Analysis",
    subtitle = "VHI Trigger Moments (Trigger: VHI ≤ 0.35 for ≥ 25 % of cropland)",
    caption = "VHI Monthly data downloaded from FAO and over FAO cropland Fraction"
  ) +
  facet_wrap(
    ~adm0_es
  )
```

## Flagging year counts
```{r}
df_vhi_yearly_flags <- df_vhi_area_thresh %>%
  group_by(adm0_es, yr = floor_date(date, "year")) %>%
  summarise(
    num_months_flagged = sum(flag_area_thresh),
    .groups = "drop"
  )

df_vhi_yearly_flags %>%
  ggplot(aes(x = yr, y = num_months_flagged, color = adm0_es)) +
  geom_point() +
  geom_line() +
  labs(
    y = "Nuber of threshold crossings",
    title = "Vegetative Health Index - Proposed Threshold Analysis",
    subtitle = "Number of VHI Trigger Moments/Year (Trigger: VHI ≤ 0.35 for ≥ 25 % of cropland)",
    caption = "VHI Monthly data downloaded from FAO and over FAO cropland Fraction"
  ) +
  scale_x_date(
    date_breaks = "1 year", date_labels = "%y"
  ) +
  theme(
    axis.text.x = element_text(angle = 90),
    axis.title.x = element_blank(),
    legend.title = element_blank()
  )
```

## % Months Breaching by both threshold

We check the % cropland pixels For each VHI threshold (0,0.05,0.1,..1) for each record/month. We then calculate the % of months where the area % threshold was greater than each threshold (0,2,4,6,..98, 100 %) for each threshold.

When we plot VHI threshold (x) by % Cropland threshold and color the fill by the % of months this combination was met we get an idea of the number/% months out of all this trigger would have been flagged.

```{r}
# define breaks and fill palette at top of each chunk so I can play with it in the different scenarios
def_breaks <- c(0, .1, seq(.2, 1, by = 0.2))
fill_cols <- RColorBrewer::brewer.pal(n = length(def_breaks), name = "YlOrRd")


df_cropland_wrangled <- df_cropland_lte_vhi_threshold %>%
  ungroup() %>%
  mutate(
    date = ym(date),
    pct_100 = pct_crop_pixels_lte_thresh * 100
  ) %>%
  rename(
    pct_crop_pixels = "pct_crop_pixels_lte_thresh"
  )


df_cropland_pct_lte <- seq(0, 1, by = 0.02) %>%
  map_dfr(
    ~ df_cropland_wrangled %>%
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = pct_crop_pixels >= area_pct_thresh
      )
  ) %>%
  group_by(
    adm0_es,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>%
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh / n(),
    .groups = "drop"
  )




df_cropland_pct_lte %>%
  mutate(
    pct_gte_thresh = ifelse(pct_gte_thresh == 0, NA, pct_gte_thresh)
  ) %>%
  ggplot(aes(x = vhi_threshold, y = area_pct_thresh)) +
  geom_tile(aes(fill = pct_gte_thresh)) +
  geom_vline(xintercept = 0.35) +
  geom_hline(yintercept = 0.25) +
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom",
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
    guide = "bins",
    breaks = def_breaks
  ) +
  labs(
    x = "VHI Threshold (Pixels ≤ threshold)",
    y = "% Cropland Pixels ≤ VHI threshold"
  ) +
  facet_wrap(~adm0_es)
```


## % years that have a breach

If we do the same process as above but this time take this time summarise to the max values per year we can view the % of years this trigger would have been flagged rather than months - this might be more similar to how people conceptualize return period

```{r}
def_breaks <- c(-Inf, seq(0, 1, by = 0.25))
fill_cols <- RColorBrewer::brewer.pal(n = length(def_breaks), name = "YlOrRd")

df_max_pct_crop_yr <- df_cropland_wrangled %>%
  group_by(adm0_es,
    threshold,
    year = year(date)
  ) %>%
  summarise(
    max_pct_crop = max(pct_crop_pixels)
  ) %>%
  ungroup()

df_cropland_max_pct_lte <- seq(0, 1, by = 0.02) %>%
  map_dfr(
    ~ df_max_pct_crop_yr %>%
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = max_pct_crop >= area_pct_thresh
      )
  ) %>%
  group_by(
    adm0_es,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>%
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh / n(),
    .groups = "drop"
  )



df_cropland_max_pct_lte %>%
  mutate(
    pct_gte_thresh = ifelse(pct_gte_thresh == 0, NA, pct_gte_thresh)
  ) %>%
  ggplot(aes(x = vhi_threshold, y = area_pct_thresh)) +
  geom_tile(aes(fill = pct_gte_thresh)) +
  geom_vline(xintercept = 0.35) +
  geom_hline(yintercept = 0.25) +
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom",
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
    guide = "bins",
    breaks = def_breaks
  ) +
  labs(
    x = "VHI Threshold (Pixels ≤ threshold)",
    y = "% Cropland Pixels ≤ VHI threshold"
  ) +
  facet_wrap(~adm0_es)
```

## How about by Season & Country

Same as above but this time by seasons of interest and country

```{r}
def_breaks <- c(-Inf, seq(0, 1, by = 0.25))
fill_cols <- RColorBrewer::brewer.pal(n = length(def_breaks), name = "YlOrRd")

df_max_pct_crop_yr_seas <- df_cropland_wrangled %>%
  mutate(
    season_vhi = case_when(

      # for primera 2 it is JJA, but we are looking at previous month
      # so it should be MJJ
      month(date) %in% c(5, 6, 7) ~ "primera 2",

      # primera 3 is SON - looking back a month ASO
      month(date) %in% c(8, 9, 10) ~ "primera 3",
      .default = NA
    )
  ) %>%
  filter(!is.na(season_vhi)) %>%
  group_by(adm0_es,
    threshold,
    year = year(date),
    season_vhi
  ) %>%
  summarise(
    max_pct_crop = max(pct_crop_pixels),
    .groups = "drop"
  ) %>%
  ungroup()

df_cropland_max_pct_season_lte <- seq(0, 1, by = 0.02) %>%
  map_dfr(
    ~ df_max_pct_crop_yr_seas %>%
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = max_pct_crop >= area_pct_thresh
      )
  ) %>%
  group_by(
    adm0_es,
    season_vhi,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>%
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh / n(),
    .groups = "drop"
  )


df_cropland_max_pct_season_lte %>%
  mutate(
    pct_gte_thresh = ifelse(pct_gte_thresh == 0, NA, pct_gte_thresh)
  ) %>%
  ggplot(aes(x = vhi_threshold, y = area_pct_thresh)) +
  geom_tile(aes(fill = pct_gte_thresh)) +
  geom_vline(xintercept = 0.35) +
  geom_hline(yintercept = 0.25) +
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom",
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
    guide = "bins",
    breaks = def_breaks
  ) +
  scale_x_continuous(labels = scales::label_number()) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "VHI Threshold (Pixels ≤ threshold)",
    y = "% Cropland Pixels ≤ VHI threshold"
  ) +
  facet_grid(cols = vars(season_vhi), rows = vars(adm0_es))
```

## What if we look at all countries together

Same as above but this time by seasons last, but this time remove the country. So here we can look at a more regional level and see the % of time the trigger/target would have been reached/flagged in any of the 4 countries

We see for "Primera 2" the proposed threshold put us in the range of ~50-60 % which roughly equates to 2 year RP which might be reasonable for this project.
```{r}
def_breaks <- c(-Inf, 0.15, 0.25, 0.35, 0.5, seq(.6, 1, by = 0.15))
fill_cols <- RColorBrewer::brewer.pal(n = length(def_breaks), name = "YlOrRd")

df_max_pct_crop_seas <- df_cropland_wrangled %>%
  mutate(
    season_vhi = case_when(

      # for primera 2 it is JJA, but we are looking at previous month
      # so it should be MJJ
      month(date) %in% c(5, 6, 7) ~ "primera 2",

      # primera 3 is SON - looking back a month ASO
      month(date) %in% c(8, 9, 10) ~ "primera 3",
      .default = NA
    )
  ) %>%
  filter(!is.na(season_vhi)) %>%
  group_by(
    threshold,
    year = year(date),
    season_vhi
  ) %>%
  summarise(
    max_pct_crop = max(pct_crop_pixels),
    .groups = "drop"
  ) %>%
  ungroup()
df_max_pct_crop_seas %>%
  print(n = 100)

df_cropland_max_pct_season_lte_together <- seq(0, 1, by = 0.02) %>%
  map_dfr(
    ~ df_max_pct_crop_seas %>%
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = max_pct_crop >= area_pct_thresh
      )
  ) %>%
  group_by(
    season_vhi,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>%
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh / n(),
    .groups = "drop"
  )


df_cropland_max_pct_season_lte_together %>%
  mutate(
    pct_gte_thresh = ifelse(pct_gte_thresh == 0, NA, pct_gte_thresh)
  ) %>%
  ggplot(aes(x = vhi_threshold, y = area_pct_thresh)) +
  geom_tile(aes(fill = pct_gte_thresh)) +
  geom_vline(xintercept = 0.35) +
  geom_hline(yintercept = 0.25) +
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom",
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
    guide = "bins",
    breaks = def_breaks
  ) +
  scale_x_continuous(labels = scales::label_number()) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "VHI Threshold (Pixels ≤ threshold)",
    y = "% Cropland Pixels ≤ VHI threshold"
  ) +
  facet_grid(cols = vars(season_vhi))
```
