---
title: "VHI"
output: html_document
date: "2023-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Visualize VHI as % of cropland per country. Will think of this in the context of the framework w/ the 2 thresholds provided by partners:

- VHI threshold= 0.35
- % Area ≤ VHI threshold (0.35) = 25%

```{r cars}
library(targets)
library(tidyverse)
library(gghdx)
gghdx()
tar_load(df_cropland_lte_vhi_threshold)

vhi_thresh <- 0.35
area_thresh <-  0.25


df_vhi_area_thresh <- df_cropland_lte_vhi_threshold %>% 
  ungroup() %>% 
  filter(round(threshold,2)== vhi_thresh) %>% 
  mutate(
    date = ym(date),
    flag_area_thresh = pct_crop_pixels_lte_thresh>=area_thresh
  )
```

# Flagging Months

```{r}

df_vhi_area_thresh %>% 
  ggplot(
    aes(x= date, y= pct_crop_pixels_lte_thresh )
  )+
  geom_point(aes(color= flag_area_thresh))+
  # geom_line()+
  geom_hline(yintercept = area_thresh,linetype="dashed",color="tomato")+
  scale_y_continuous_hdx(labels=scales::label_percent())+
    labs(
    y="% Cropland",
       title = "Vegetative Health Index - Proposed Threshold Analysis",
       subtitle = "VHI Trigger Moments (Trigger: VHI ≤ 0.35 for ≥ 25 % of cropland)",
       caption = "VHI Monthly data downloaded from FAO and over FAO cropland Fraction"
  )+
  facet_wrap(
    ~adm0_es
  )

```

## Flagging year counts
```{r}
df_vhi_yearly_flags <- df_vhi_area_thresh %>% 
  group_by(adm0_es,yr= floor_date(date, "year")) %>% 
  summarise(
    num_months_flagged  =sum(flag_area_thresh),
    .groups="drop"
  )

df_vhi_yearly_flags %>%
  ggplot(aes(x= yr, y= num_months_flagged,color=adm0_es) )+
  geom_point()+
  geom_line()+
  labs(
    y="Nuber of threshold crossings",
       title = "Vegetative Health Index - Proposed Threshold Analysis",
       subtitle ="Number of VHI Trigger Moments/Year (Trigger: VHI ≤ 0.35 for ≥ 25 % of cropland)",
       caption = "VHI Monthly data downloaded from FAO and over FAO cropland Fraction"
  )+
  
  scale_x_date(
    date_breaks = "1 year",date_labels = "%y"
  )+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.x = element_blank(),
    legend.title = element_blank()
    
  )
```

## % Months Breaching by both threshold

```{r}
df_cropland_wrangled <- df_cropland_lte_vhi_threshold %>% 
  ungroup() %>% 
  mutate(
    date = ym(date),
    pct_100 = pct_crop_pixels_lte_thresh*100
  ) %>%
  rename(
    pct_crop_pixels ="pct_crop_pixels_lte_thresh"
  ) 


df_cropland_pct_lte <- seq(0,1, by=0.02) %>% 
  map_dfr(
    ~df_cropland_wrangled %>% 
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = pct_crop_pixels>=area_pct_thresh
      )
  ) %>% 
  group_by(
    adm0_es,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>% 
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh/n(),
    .groups = "drop"
  ) 
  
  
def_breaks <- c(0,.1,seq(.2,1 ,by=0.2))
fill_cols<- RColorBrewer::brewer.pal(n = length(def_breaks),name = "Blues")
  
df_cropland_pct_lte %>% 
    mutate(
    pct_gte_thresh= ifelse(pct_gte_thresh==0,NA,pct_gte_thresh)
  ) %>% 
  ggplot(aes(x=vhi_threshold, y= area_pct_thresh))+
  geom_tile(aes(fill=pct_gte_thresh))+
  geom_vline(xintercept = 0.35)+
  geom_hline(yintercept = 0.25)+
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom", 
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
               guide = "bins",
               breaks = def_breaks
    
      )+
  labs(
    x= "VHI Threshold (Pixels ≤ threshold)",
    y ="% Cropland Pixels ≤ VHI threshold"
  )+
  facet_wrap(~adm0_es)
```


## % years that have a breach
by both thresholds
```{r}
df_max_pct_crop_yr <- df_cropland_wrangled %>% 
      group_by(adm0_es,
               threshold,year=year(date)) %>% 
      summarise(
        max_pct_crop = max(pct_crop_pixels)
      ) %>% 
  ungroup()

df_cropland_max_pct_lte <- seq(0,1, by=0.02) %>% 
  map_dfr(
    ~ df_max_pct_crop_yr%>% 
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = max_pct_crop>=area_pct_thresh
      )
  ) %>% 
  group_by(
    adm0_es,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>% 
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh/n(),
    .groups = "drop"
  )



df_cropland_max_pct_lte %>% 
    mutate(
    pct_gte_thresh= ifelse(pct_gte_thresh==0,NA,pct_gte_thresh)
  ) %>% 
  ggplot(aes(x=vhi_threshold, y= area_pct_thresh))+
  geom_tile(aes(fill=pct_gte_thresh))+
  geom_vline(xintercept = 0.35)+
  geom_hline(yintercept = 0.25)+
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom", 
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
               guide = "bins",
               breaks = def_breaks
    
      )+
  labs(
    x= "VHI Threshold (Pixels ≤ threshold)",
    y ="% Cropland Pixels ≤ VHI threshold"
  )+
  facet_wrap(~adm0_es)
```

## How about by Season & Country

Here we see for each country the % of time the different threshold combinations are met

```{r}
df_max_pct_crop_yr_seas <- df_cropland_wrangled %>% 
  mutate(
    season_vhi = case_when(
      
      # for primera 2 it is JJA, but we are looking at previous month
      # so it should be MJJ
      month(date) %in% c(5,6,7) ~ "primera 2",
      
      # primera 3 is SON - looking back a month ASO
      month(date) %in% c(8,9,10) ~ "primera 3",
      .default = NA
    )
  ) %>% 
  filter(!is.na(season_vhi)) %>% 
      group_by(adm0_es,
               threshold,
               year=year(date),
               season_vhi
               ) %>% 
      summarise(
        max_pct_crop = max(pct_crop_pixels),
        .groups = "drop"
        
      ) %>% 
  ungroup()

df_cropland_max_pct_season_lte<- seq(0,1, by=0.02) %>% 
  map_dfr(
    ~ df_max_pct_crop_yr_seas %>% 
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = max_pct_crop>=area_pct_thresh
      )
  ) %>% 
  group_by(
    adm0_es,
    season_vhi,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>% 
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh/n(),
    .groups = "drop"
  )

def_breaks <- c(-Inf,seq(0,1, by=0.25))
fill_cols<- RColorBrewer::brewer.pal(n = length(def_breaks),name = "Blues")

df_cropland_max_pct_season_lte %>% 
  mutate(
    pct_gte_thresh= ifelse(pct_gte_thresh==0,NA,pct_gte_thresh)
  ) %>% 
  ggplot(aes(x=vhi_threshold, y= area_pct_thresh))+
  geom_tile(aes(fill=pct_gte_thresh))+
  geom_vline(xintercept = 0.35)+
  geom_hline(yintercept = 0.25)+
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom", 
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
               guide = "bins",
               breaks = def_breaks
    
      )+
  labs(
    x= "VHI Threshold (Pixels ≤ threshold)",
    y ="% Cropland Pixels ≤ VHI threshold"
  )+
  facet_grid(cols= vars(season_vhi), rows=vars(adm0_es))


```

## What if we look at all countries together

If we combine all countries we can see it's met betwee 50-75 % of the time in primera one and 

```{r}
def_breaks <- c(-Inf,0.15,0.25,0.35,0.5,seq(.6,1, by=0.15))
fill_cols<- RColorBrewer::brewer.pal(n = length(def_breaks),name = "YlOrRd")

df_max_pct_crop_seas <- df_cropland_wrangled %>% 
  mutate(
    season_vhi = case_when(
      
      # for primera 2 it is JJA, but we are looking at previous month
      # so it should be MJJ
      month(date) %in% c(5,6,7) ~ "primera 2",
      
      # primera 3 is SON - looking back a month ASO
      month(date) %in% c(8,9,10) ~ "primera 3",
      .default = NA
    )
  ) %>% 
  filter(!is.na(season_vhi)) %>% 
      group_by(
               threshold,
               year=year(date),
               season_vhi
               ) %>% 
      summarise(
        max_pct_crop = max(pct_crop_pixels),
        .groups = "drop"
        
      ) %>% 
  ungroup()
df_max_pct_crop_seas %>% 
  print(n=100)

df_cropland_max_pct_season_lte_together<- seq(0,1, by=0.02) %>% 
  map_dfr(
    ~ df_max_pct_crop_seas %>% 
      mutate(
        area_pct_thresh = .x,
        gte_area_pct_thresh = max_pct_crop>=area_pct_thresh
      )
  ) %>% 
  group_by(
    season_vhi,
    vhi_threshold = threshold,
    area_pct_thresh
  ) %>% 
  summarise(
    n_gte_thresh = sum(gte_area_pct_thresh),
    pct_gte_thresh = n_gte_thresh/n(),
    .groups = "drop"
  )


df_cropland_max_pct_season_lte_together %>% 
  mutate(
    pct_gte_thresh= ifelse(pct_gte_thresh==0,NA,pct_gte_thresh)
  ) %>% 
  ggplot(aes(x=vhi_threshold, y= area_pct_thresh))+
  geom_tile(aes(fill=pct_gte_thresh))+
  geom_vline(xintercept = 0.35)+
  geom_hline(yintercept = 0.25)+
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom", 
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
               guide = "bins",
               breaks = def_breaks
    
      )+
  labs(
    x= "VHI Threshold (Pixels ≤ threshold)",
    y ="% Cropland Pixels ≤ VHI threshold"
  )+
  facet_grid(cols= vars(season_vhi))


```