---
title: "AOI Sensitivity Exploration"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(lubridate)
library(sf)
library(cumulus)
library(yardstick)
library(gghdx)
library(gt)
gghdx()
```

## Overview

Chapter 07 established SEAS5 skill using the original framework AOI:

- **Guatemala**: Chiquimula (GT20)
- **Honduras**: El Paraíso + Francisco Morazán (HN07/HN08).
- **El Salvador**: Ahuachapán + Santa Ana (SV01/SV12)

This chapter explores two dimensions of geographic sensitivity:

1. **Admin expansion**: Does adding neighbouring admin units (GT16 Zacapa, SV13 Morazán, SV05 La Libertad) change skill?
2. **Spatial scale**: How does country-level aggregation compare to sub-national?

## Data Preparation

```{r}
#| label: load-data
#| cache: true

PRIMERA_MONTHS <- 5:8
POSTRERA_MONTHS <- 9:11
PRIMERA_ISSUED_MONTHS <- c(3, 4, 5)
POSTRERA_ISSUED_MONTHS <- c(6, 7, 8, 9)

# All candidate admin-1 pcodes
CANDIDATE_PCODES <- c(
  "GT20", "GT16",          # Guatemala: Chiquimula, Alta Verapaz
  "GT19", "GT02",           # Guatemala: Zacapa, El Progreso
  "GT21",                  # Guatemala: Jalapa
  "GT14", "GT15",          # Guatemala: Quiché, Baja Verapaz
  "HN07", "HN08",          # Honduras: El Paraíso, Francisco Morazán
  "SV01", "SV12",          # El Salvador confirmed: Ahuachapán, Santa Ana
  "SV13", "SV05"           # El Salvador candidates: Morazán, La Libertad
)

con <- pg_con()

df_weights <- tbl(con, "polygon") |>
  mutate(across(pcode, as.character)) |>
  filter(adm_level == 1, pcode %in% CANDIDATE_PCODES) |>
  select(pcode, iso3, name, seas5_n_upsampled_pixels) |>
  collect()

# Admin-1 level data
df_seas5_raw <- tbl(con, "seas5") |>
  mutate(across(pcode, as.character)) |>
  filter(adm_level == 1, pcode %in% CANDIDATE_PCODES) |>
  collect()

df_era5_raw <- tbl(con, "era5") |>
  mutate(across(pcode, as.character)) |>
  filter(pcode %in% CANDIDATE_PCODES) |>
  collect()

# Country-level (adm0) data
df_seas5_country <- tbl(con, "seas5") |>
  mutate(across(pcode, as.character)) |>
  filter(adm_level == 0, iso3 %in% c("GTM", "HND", "SLV")) |>
  collect()

df_era5_country <- tbl(con, "era5") |>
  mutate(across(pcode, as.character)) |>
  filter(adm_level == 0, iso3 %in% c("GTM", "HND", "SLV")) |>
  collect()

DBI::dbDisconnect(con)
```

```{r}
#| label: process-admin-data

# Process admin-1 to seasonal totals
df_seas5_mm <- df_seas5_raw |>
  mutate(value_mm = days_in_month(valid_date) * mean)

df_seas5_seasonal <- bind_rows(
  seas5_aggregate_forecast(df_seas5_mm, value = "value_mm", valid_months = PRIMERA_MONTHS,
                           by = c("iso3", "pcode", "issued_date")) |> mutate(window = "primera"),
  seas5_aggregate_forecast(df_seas5_mm, value = "value_mm", valid_months = POSTRERA_MONTHS,
                           by = c("iso3", "pcode", "issued_date")) |> mutate(window = "postrera")
) |>
  rename(fcst_mm = value_mm) |>
  mutate(
    year = year(issued_date),
    issued_month = month(issued_date)
  ) |>
  filter(
    (window == "primera" & issued_month %in% PRIMERA_ISSUED_MONTHS) |
    (window == "postrera" & issued_month %in% POSTRERA_ISSUED_MONTHS)
  )

df_era5_monthly <- df_era5_raw |>
  mutate(
    year = year(valid_date),
    month = month(valid_date),
    value_mm = mean * days_in_month(valid_date)
  )

df_era5_seasonal <- bind_rows(
  df_era5_monthly |>
    filter(month %in% PRIMERA_MONTHS) |>
    group_by(pcode, iso3, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "primera"),
  df_era5_monthly |>
    filter(month %in% POSTRERA_MONTHS) |>
    group_by(pcode, iso3, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "postrera")
)
```

```{r}
#| label: process-country-data

# Process country-level to seasonal totals
df_seas5_country_mm <- df_seas5_country |>
  mutate(value_mm = days_in_month(valid_date) * mean)

df_seas5_country_seasonal <- bind_rows(
  seas5_aggregate_forecast(df_seas5_country_mm, value = "value_mm", valid_months = PRIMERA_MONTHS,
                           by = c("iso3", "pcode", "issued_date")) |> mutate(window = "primera"),
  seas5_aggregate_forecast(df_seas5_country_mm, value = "value_mm", valid_months = POSTRERA_MONTHS,
                           by = c("iso3", "pcode", "issued_date")) |> mutate(window = "postrera")
) |>
  rename(fcst_mm = value_mm) |>
  mutate(
    year = year(issued_date),
    issued_month = month(issued_date)
  ) |>
  filter(
    (window == "primera" & issued_month %in% PRIMERA_ISSUED_MONTHS) |
    (window == "postrera" & issued_month %in% POSTRERA_ISSUED_MONTHS)
  )

df_era5_country_monthly <- df_era5_country |>
  mutate(
    year = year(valid_date),
    month = month(valid_date),
    value_mm = mean * days_in_month(valid_date)
  )

df_era5_country_seasonal <- bind_rows(
  df_era5_country_monthly |>
    filter(month %in% PRIMERA_MONTHS) |>
    group_by(pcode, iso3, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "primera"),
  df_era5_country_monthly |>
    filter(month %in% POSTRERA_MONTHS) |>
    group_by(pcode, iso3, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "postrera")
)
```

```{r}
#| label: skill-functions

calc_rp_threshold <- function(x, rp_target = 4, direction = -1) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 3) return(NA_real_)
  ranks <- rank(x * -direction, ties.method = "average")
  rp <- (n + 1) / ranks
  approx(rp, x, xout = rp_target, rule = 2)$y
}

calc_skill_metrics <- function(df, baseline_start = 1981, baseline_end = 2024, rp_target = 4) {
  df_baseline <- df |>
    filter(year >= baseline_start, year <= baseline_end)

  obs_thresholds <- df_baseline |>
    group_by(aoi_config, window) |>
    summarise(obs_thresh = calc_rp_threshold(obs_mm, rp_target, -1), .groups = "drop")

  fcst_thresholds <- df_baseline |>
    group_by(aoi_config, window, leadtime) |>
    summarise(fcst_thresh = calc_rp_threshold(fcst_mm, rp_target, -1), .groups = "drop")

  drought_levels <- c("drought", "no_drought")

  df_baseline |>
    left_join(obs_thresholds, by = c("aoi_config", "window")) |>
    left_join(fcst_thresholds, by = c("aoi_config", "window", "leadtime")) |>
    mutate(
      truth = fct(
        if_else(obs_mm <= obs_thresh, "drought", "no_drought"),
        levels = drought_levels
      ),
      estimate = fct(
        if_else(fcst_mm <= fcst_thresh, "drought", "no_drought"),
        levels = drought_levels
      )
    ) |>
    filter(!is.na(truth), !is.na(estimate)) |>
    group_by(aoi_config, window, leadtime) |>
    summarise(
      n_years = n_distinct(year),
      n_drought = sum(truth == "drought"),
      spearman = cor(fcst_mm, obs_mm, method = "spearman", use = "complete.obs"),
      roc_auc = roc_auc_vec(truth, -fcst_mm, event_level = "first"),
      f1 = f_meas_vec(truth, estimate, event_level = "first"),
      .groups = "drop"
    )
}
```

## Rainfall Coherence Map

Before examining forecast skill across AOI configurations, we first check how well seasonal rainfall in each Guatemala department correlates with Chiquimula (GT20) — the current framework AOI. Departments with low correlation experience droughts at different times, making them poor candidates for combining into a single trigger.

```{r}
#| label: load-gtm-spatial
#| cache: true

# Get all GTM admin-1 ERA5 data
con <- pg_con()

df_era5_gtm_all <- tbl(con, "era5") |>
  mutate(across(pcode, as.character)) |>
  filter(iso3 == "GTM", adm_level == 1) |>
  collect()

DBI::dbDisconnect(con)

# Get admin-1 boundaries
gdf_gtm <- cumulus::download_fieldmaps_sf(iso3 = "gtm", layer = "gtm_adm1")$gtm_adm1
```

```{r}
#| label: calc-correlation-map

# Aggregate to seasonal totals
df_gtm_monthly <- df_era5_gtm_all |>
  mutate(
    year = year(valid_date),
    month = month(valid_date),
    value_mm = mean * days_in_month(valid_date)
  )

df_gtm_seasonal <- bind_rows(
  df_gtm_monthly |>
    filter(month %in% PRIMERA_MONTHS) |>
    group_by(pcode, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "primera"),
  df_gtm_monthly |>
    filter(month %in% POSTRERA_MONTHS) |>
    group_by(pcode, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "postrera")
) |>
  filter(year >= 1981)

# Calculate Spearman correlation of each admin with GT20
gt20_obs <- df_gtm_seasonal |>
  filter(pcode == "GT20") |>
  select(year, window, obs_mm_gt20 = obs_mm)

df_cor_gt20 <- df_gtm_seasonal |>
  left_join(gt20_obs, by = c("year", "window")) |>
  group_by(pcode, window) |>
  summarise(
    spearman = cor(obs_mm, obs_mm_gt20, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  )
```

```{r}
#| label: correlation-map
#| fig-width: 10
#| fig-height: 6

# Join correlation to spatial data
# Match on pcode - fieldmaps may use different column name
pcode_col <- intersect(c("ADM1_PCODE", "adm1_pcode", "pcode"), names(gdf_gtm))[1]
gdf_gtm_cor <- gdf_gtm |>
  rename(pcode = !!pcode_col) |>
  left_join(df_cor_gt20, by = "pcode")

gdf_gtm_cor |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    )
  ) |>
  ggplot() +
  geom_sf(aes(fill = spearman), color = "white", linewidth = 0.3) +
  geom_sf(
    data = gdf_gtm_cor |> filter(pcode == "GT20") |> head(1),
    fill = NA, color = "black", linewidth = 1.5
  ) +
  geom_sf_text(aes(label = round(spearman, 2)), size = 2.5, color = "black") +
  facet_wrap(~window_label) +
  scale_fill_gradient2(
    low = "#D73027", mid = "#FFFFBF", high = "#1A9850",
    midpoint = 0.75, limits = c(0.4, 1), name = "Spearman r\nvs Chiquimula",
    oob = scales::squish
  ) +
  labs(
    title = "Rainfall Correlation with Chiquimula (GT20)",
    subtitle = "ERA5 seasonal totals | 1981-2024 | Black border = Chiquimula"
  ) +
  theme_void() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right",
    plot.title = element_text(size = 14, face = "bold")
  )
```

## Admin Expansion

How does skill change when we expand the geographic footprint within each country?

```{r}
#| label: build-aoi-configs

# Build pcode-to-name lookup from polygon table (all admin-1, not just those with SEAS5 data)
df_admin_names <- tbl(pg_con(), "polygon") |>
  mutate(across(pcode, as.character)) |>
  filter(adm_level == 1, pcode %in% CANDIDATE_PCODES) |>
  select(pcode, name) |>
  collect()
pcode_to_name <- setNames(df_admin_names$name, df_admin_names$pcode)

relabel_config <- function(config_name) {
  # "GTM: GT20+GT16" → "GTM: Chiquimula + Alta Verapaz"
  parts <- str_split(config_name, ": ", n = 2)[[1]]
  prefix <- parts[1]
  pcodes <- str_split(parts[2], "\\+")[[1]]
  names <- ifelse(pcodes %in% names(pcode_to_name), pcode_to_name[pcodes], pcodes)
  paste0(prefix, ": ", paste(names, collapse = " + "))
}

# Define AOI configurations to compare
aoi_configs <- list(
  # Guatemala
  "GTM: GT20" = c("GT20"),
  "GTM: GT20+GT16" = c("GT20", "GT16"),
  "GTM: GT20+GT19" = c("GT20", "GT19"),
  "GTM: GT20+GT02" = c("GT20", "GT02"),
  "GTM: GT20+GT21+GT02+GT19" = c("GT20", "GT21", "GT02", "GT19"),
  "GTM: GT20+GT21+GT02" = c("GT20", "GT21", "GT02"),
  "GTM: GT16" = c("GT16"),
  "GTM: GT15" = c("GT15"),
  "GTM: GT14" = c("GT14"),
  "GTM: GT14+GT15" = c("GT14", "GT15"),
  # Honduras
  "HND: HN07+HN08" = c("HN07", "HN08"),
  # El Salvador
  "SLV: SV01+SV12" = c("SV01", "SV12"),
  "SLV: SV01+SV12+SV13" = c("SV01", "SV12", "SV13"),
  "SLV: SV01+SV12+SV13+SV05" = c("SV01", "SV12", "SV13", "SV05")
)

# Relabel config names to use admin names
names(aoi_configs) <- map_chr(names(aoi_configs), relabel_config)

# Build weighted-mean aggregated datasets for each config
aggregate_config <- function(pcodes, config_name) {
  df_fcst <- df_seas5_seasonal |>
    filter(pcode %in% pcodes) |>
    left_join(df_weights |> select(pcode, seas5_n_upsampled_pixels), by = "pcode")

  df_obs <- df_era5_seasonal |>
    filter(pcode %in% pcodes) |>
    left_join(df_weights |> select(pcode, seas5_n_upsampled_pixels), by = "pcode")

  df_joined <- df_fcst |>
    left_join(
      df_obs |> select(pcode, year, window, obs_mm),
      by = c("pcode", "year", "window")
    ) |>
    filter(!is.na(obs_mm)) |>
    group_by(year, window, leadtime, issued_date) |>
    summarise(
      fcst_mm = weighted.mean(fcst_mm, w = seas5_n_upsampled_pixels),
      obs_mm = weighted.mean(obs_mm, w = seas5_n_upsampled_pixels),
      .groups = "drop"
    ) |>
    mutate(aoi_config = config_name)

  df_joined
}

df_all_configs <- imap_dfr(aoi_configs, ~aggregate_config(.x, .y))

# Add country-level configs
df_country_joined <- df_seas5_country_seasonal |>
  left_join(
    df_era5_country_seasonal |> select(pcode, year, window, obs_mm),
    by = c("pcode", "year", "window")
  ) |>
  filter(!is.na(obs_mm)) |>
  mutate(
    aoi_config = case_when(
      iso3 == "GTM" ~ "GTM: Country",
      iso3 == "HND" ~ "HND: Country",
      iso3 == "SLV" ~ "SLV: Country"
    )
  )

df_all_configs <- bind_rows(df_all_configs, df_country_joined)
```

```{r}
#| label: admin-skill-calc

df_skill_admin <- calc_skill_metrics(df_all_configs)

# Shared theme for faceted horizontal bar charts
theme_facet_bars <- theme(
  legend.position = "none",
  panel.spacing = unit(0.8, "lines"),
  strip.background = element_rect(fill = "grey90", color = "grey60"),
  strip.text = element_text(face = "bold", size = 11),
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4)
)
```

### ROC-AUC Comparison

::: {.panel-tabset}

#### Guatemala

```{r}
#| label: gtm-admin-comparison
#| fig-width: 12
#| fig-height: 10

df_skill_admin |>
  filter(str_starts(aoi_config, "GTM")) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime),
    aoi_label = str_remove(aoi_config, "^GTM: ")
  ) |>
  ggplot(aes(x = roc_auc, y = reorder(aoi_label, roc_auc), fill = aoi_label)) +
  geom_col() +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black", linewidth = 0.6) +
  facet_grid(lt_label ~ window_label) +
  scale_x_continuous(breaks = seq(0.4, 1, 0.1)) +
  coord_cartesian(xlim = c(0.4, 1)) +
  labs(
    title = "Guatemala: AOI Sensitivity",
    subtitle = "ROC-AUC for RP4 drought | 1981-2024 baseline",
    x = "ROC-AUC", y = NULL
  ) +
  theme_facet_bars
```

#### Honduras

```{r}
#| label: hnd-admin-comparison
#| fig-width: 10
#| fig-height: 8

df_skill_admin |>
  filter(str_starts(aoi_config, "HND")) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime),
    aoi_label = str_remove(aoi_config, "^HND: ")
  ) |>
  ggplot(aes(x = roc_auc, y = reorder(aoi_label, roc_auc), fill = aoi_label)) +
  geom_col() +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black", linewidth = 0.6) +
  facet_grid(lt_label ~ window_label) +
  scale_x_continuous(breaks = seq(0.4, 1, 0.1)) +
  coord_cartesian(xlim = c(0.4, 1)) +
  labs(
    title = "Honduras: AOI Sensitivity",
    subtitle = "ROC-AUC for RP4 drought | 1981-2024 baseline",
    x = "ROC-AUC", y = NULL
  ) +
  theme_facet_bars
```

#### El Salvador

```{r}
#| label: slv-admin-comparison
#| fig-width: 10
#| fig-height: 8

df_skill_admin |>
  filter(str_starts(aoi_config, "SLV")) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime),
    aoi_label = str_remove(aoi_config, "^SLV: ")
  ) |>
  ggplot(aes(x = roc_auc, y = reorder(aoi_label, roc_auc), fill = aoi_label)) +
  geom_col() +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black", linewidth = 0.6) +
  facet_grid(lt_label ~ window_label) +
  scale_x_continuous(breaks = seq(0.4, 1, 0.1)) +
  coord_cartesian(xlim = c(0.4, 1)) +
  labs(
    title = "El Salvador: AOI Sensitivity",
    subtitle = "ROC-AUC for RP4 drought | 1981-2024 baseline",
    x = "ROC-AUC", y = NULL
  ) +
  theme_facet_bars
```

:::

### F1 Score Comparison

```{r}
#| label: admin-f1-heatmap
#| fig-width: 14
#| fig-height: 8

df_skill_admin |>
  filter(
    (window == "primera" & leadtime %in% 0:2) |
    (window == "postrera" & leadtime %in% 0:3)
  ) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera", "Postrera")
    )
  ) |>
  ggplot(aes(x = factor(leadtime), y = aoi_config, fill = f1)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", f1)), size = 4, fontface = "bold") +
  facet_wrap(~window_label, scales = "free_x") +
  scale_fill_gradient2(
    low = "#D73027", mid = "#FFFFBF", high = "#1A9850",
    midpoint = 0.4, limits = c(0, 0.8), name = "F1 Score",
    oob = scales::squish, na.value = "grey80"
  ) +
  labs(
    title = "F1 Score Across AOI Configurations",
    subtitle = "RP4 drought | 1981-2024 baseline",
    x = "Leadtime (months)", y = NULL
  ) +
  theme(panel.grid = element_blank())
```

### Spearman Correlation

```{r}
#| label: admin-spearman-heatmap
#| fig-width: 14
#| fig-height: 8

df_skill_admin |>
  filter(
    (window == "primera" & leadtime %in% 0:2) |
    (window == "postrera" & leadtime %in% 0:3)
  ) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera", "Postrera")
    )
  ) |>
  ggplot(aes(x = factor(leadtime), y = aoi_config, fill = spearman)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", spearman)), size = 4, fontface = "bold") +
  facet_wrap(~window_label, scales = "free_x") +
  scale_fill_gradient2(
    low = "#D73027", mid = "#FFFFBF", high = "#1A9850",
    midpoint = 0.4, limits = c(0, 0.9), name = "Spearman r",
    oob = scales::squish, na.value = "grey80"
  ) +
  labs(
    title = "Spearman Correlation Across AOI Configurations",
    subtitle = "Continuous skill metric | 1981-2024 baseline",
    x = "Leadtime (months)", y = NULL
  ) +
  theme(panel.grid = element_blank())
```

### Summary Table

```{r}
#| label: admin-summary-table

df_skill_admin |>
  filter(
    (window == "primera" & leadtime %in% 0:2) |
    (window == "postrera" & leadtime %in% 0:1)
  ) |>
  mutate(
    score = sprintf("%.2f (%.2f)", roc_auc, f1),
    lt_label = paste0("LT", leadtime)
  ) |>
  select(aoi_config, window, lt_label, score) |>
  pivot_wider(names_from = lt_label, values_from = score) |>
  arrange(window, aoi_config) |>
  gt(groupname_col = "window") |>
  tab_header(
    title = md("**Admin Expansion: Skill Summary**"),
    subtitle = "Format: ROC-AUC (F1) | RP4 drought | 1981-2024 baseline"
  ) |>
  cols_label(aoi_config = "AOI Configuration")
```

### Full Summary Table

```{r}
#| label: full-summary-table

df_skill_admin |>
  filter(
    (window == "primera" & leadtime %in% 0:2) |
    (window == "postrera" & leadtime %in% 0:1)
  ) |>
  mutate(
    score = sprintf("%.2f (%.2f)", roc_auc, f1),
    lt_label = paste0("LT", leadtime)
  ) |>
  select(aoi_config, window, lt_label, score) |>
  pivot_wider(names_from = lt_label, values_from = score) |>
  arrange(window, aoi_config) |>
  gt(groupname_col = "window") |>
  tab_header(
    title = md("**AOI Sensitivity: Skill Summary**"),
    subtitle = "Format: ROC-AUC (F1) | RP4 drought | 1981-2024 baseline"
  ) |>
  cols_label(aoi_config = "AOI Configuration")
```

## RP3 vs RP4 Drought Definition

How sensitive are the results to the drought definition? Here we compare RP4 (1 in 4 year, ~25% of years) with RP3 (1 in 3 year, ~33% of years).

```{r}
#| label: rp3-skill-calc

df_skill_rp3 <- calc_skill_metrics(df_all_configs, rp_target = 3) |> mutate(rp = "RP3")
df_skill_rp4 <- df_skill_admin |> mutate(rp = "RP4")
df_skill_rp_comparison <- bind_rows(df_skill_rp3, df_skill_rp4)
```

::: {.panel-tabset}

### Guatemala

```{r}
#| label: gtm-rp-comparison
#| fig-width: 12
#| fig-height: 10

df_skill_rp_comparison |>
  filter(str_starts(aoi_config, "GTM"),
         (window == "primera" & leadtime %in% 0:2) |
         (window == "postrera" & leadtime %in% 0:2)) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime),
    aoi_label = str_remove(aoi_config, "^GTM: ")
  ) |>
  ggplot(aes(x = roc_auc, y = reorder(aoi_label, roc_auc), fill = aoi_label, alpha = rp)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black", linewidth = 0.6) +
  facet_grid(lt_label ~ window_label) +
  scale_alpha_manual(values = c("RP3" = 0.5, "RP4" = 1)) +
  scale_x_continuous(breaks = seq(0.4, 1, 0.1)) +
  coord_cartesian(xlim = c(0.4, 1)) +
  labs(
    title = "Guatemala: RP3 vs RP4 Drought Definition",
    subtitle = "Solid = RP4, transparent = RP3 | 1981-2024 baseline",
    x = "ROC-AUC", y = NULL, alpha = "Threshold"
  ) +
  theme_facet_bars +
  theme(legend.position = "bottom") +
  guides(fill = "none", alpha = guide_legend(nrow = 1))
```

### Honduras

```{r}
#| label: hnd-rp-comparison
#| fig-width: 10
#| fig-height: 8

df_skill_rp_comparison |>
  filter(str_starts(aoi_config, "HND"),
         (window == "primera" & leadtime %in% 0:2) |
         (window == "postrera" & leadtime %in% 0:2)) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime),
    aoi_label = str_remove(aoi_config, "^HND: ")
  ) |>
  ggplot(aes(x = roc_auc, y = reorder(aoi_label, roc_auc), fill = aoi_label, alpha = rp)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black", linewidth = 0.6) +
  facet_grid(lt_label ~ window_label) +
  scale_alpha_manual(values = c("RP3" = 0.5, "RP4" = 1)) +
  scale_x_continuous(breaks = seq(0.4, 1, 0.1)) +
  coord_cartesian(xlim = c(0.4, 1)) +
  labs(
    title = "Honduras: RP3 vs RP4 Drought Definition",
    subtitle = "Solid = RP4, transparent = RP3 | 1981-2024 baseline",
    x = "ROC-AUC", y = NULL, alpha = "Threshold"
  ) +
  theme_facet_bars +
  theme(legend.position = "bottom") +
  guides(fill = "none", alpha = guide_legend(nrow = 1))
```

### El Salvador

```{r}
#| label: slv-rp-comparison
#| fig-width: 10
#| fig-height: 8

df_skill_rp_comparison |>
  filter(str_starts(aoi_config, "SLV"),
         (window == "primera" & leadtime %in% 0:2) |
         (window == "postrera" & leadtime %in% 0:2)) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime),
    aoi_label = str_remove(aoi_config, "^SLV: ")
  ) |>
  ggplot(aes(x = roc_auc, y = reorder(aoi_label, roc_auc), fill = aoi_label, alpha = rp)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black", linewidth = 0.6) +
  facet_grid(lt_label ~ window_label) +
  scale_alpha_manual(values = c("RP3" = 0.5, "RP4" = 1)) +
  scale_x_continuous(breaks = seq(0.4, 1, 0.1)) +
  coord_cartesian(xlim = c(0.4, 1)) +
  labs(
    title = "El Salvador: RP3 vs RP4 Drought Definition",
    subtitle = "Solid = RP4, transparent = RP3 | 1981-2024 baseline",
    x = "ROC-AUC", y = NULL, alpha = "Threshold"
  ) +
  theme_facet_bars +
  theme(legend.position = "bottom") +
  guides(fill = "none", alpha = guide_legend(nrow = 1))
```

:::

## Observations

*To be filled in after running the analysis.*

## Start Network: Baja Verapaz & Quiché {#sec-start-network}

::: {.callout-note}
## Partner-specific analysis

The Start Network is exploring a separate trigger for **Baja Verapaz (GT15)** and **Quiché (GT14)**. This section summarises the AOI skill assessment and recommended monitoring configuration. Detailed admin 2 decomposition is in @sec-sn-appendix.
:::

The Guatemala bar charts above already include Quiché, Baja Verapaz, and their combination (GT14+GT15). At the RP4 drought definition, the merged GT14+GT15 AOI exceeds ROC-AUC 0.7 at primera LT0 and LT1, and performs well for postrera. The main weak point is primera LT2, where the merged skill is dragged down by low-skill zones in northern Quiché. However, Baja Verapaz alone retains adequate skill at LT2.

### Recommendation

For the Start Network trigger covering Baja Verapaz and Quiché:

- **Primera LT0-LT1 and all postrera lead times**: Use the merged **GT14+GT15** AOI. The combined area has ROC-AUC above 0.7 at these lead times, and spatial averaging benefits skill.
- **Primera LT2**: Monitor **Baja Verapaz (GT15) only**. The merged AOI falls below acceptable skill at this lead time due to poorly-predicted zones in northern Quiché. If Baja Verapaz activates at primera LT2, the Start Network can still decide how best to prioritise resources between the two departments.

## Appendix: Admin 2 Skill Decomposition for GT14+GT15 {#sec-sn-appendix}

This appendix contains supplementary spatial analysis supporting the Start Network recommendation above, including inter-departmental rainfall correlation, admin 2-level forecast skill maps, leave-one-out decomposition, and forecast-similarity clustering.

### Correlation Matrix

```{r}
#| label: fig-gtm-corrplot
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Spearman rank correlation of ERA5 seasonal rainfall across all Guatemala admin 1 departments (1981-2024)."

library(corrplot)

# Get department names from spatial data
name_col <- intersect(c("ADM1_ES", "adm1_es", "ADM1_EN", "adm1_en", "shapeName"), names(gdf_gtm))[1]
gtm_names <- gdf_gtm |>
  st_drop_geometry() |>
  rename(pcode = !!pcode_col, dept_name = !!name_col) |>
  select(pcode, dept_name) |>
  distinct()

for (szn in c("primera", "postrera")) {
  era5_wide <- df_gtm_seasonal |>
    filter(window == szn) |>
    left_join(gtm_names, by = "pcode") |>
    mutate(label = paste0(dept_name, " (", pcode, ")")) |>
    select(year, label, obs_mm) |>
    pivot_wider(names_from = label, values_from = obs_mm)

  cor_mat <- cor(era5_wide |> select(-year), method = "spearman", use = "complete.obs")

  corrplot(cor_mat, method = "color", type = "lower",
           tl.col = "black", tl.cex = 0.55, tl.srt = 45,
           addCoef.col = "black", number.cex = 0.45,
           cl.cex = 0.7, mar = c(0, 0, 2, 0),
           title = paste0("ERA5 ", tools::toTitleCase(szn),
                          " Rainfall: Spearman Correlation (1981-2024)"))
}
```

### Admin 2 Forecast Skill Map

```{r}
#| label: load-gtm-adm2
#| cache: true

con <- pg_con()

df_seas5_gtm_adm2 <- tbl(con, "seas5") |>
  mutate(across(pcode, as.character)) |>
  filter(iso3 == "GTM", adm_level == 2) |>
  collect()

df_era5_gtm_adm2 <- tbl(con, "era5") |>
  mutate(across(pcode, as.character)) |>
  filter(iso3 == "GTM", adm_level == 2) |>
  collect()

DBI::dbDisconnect(con)

gdf_gtm_adm2 <- cumulus::download_fieldmaps_sf(iso3 = "gtm", layer = "gtm_adm2")$gtm_adm2
```

```{r}
#| label: calc-gtm-adm2-skill

# Process to seasonal totals
df_seas5_adm2_mm <- df_seas5_gtm_adm2 |>
  mutate(value_mm = days_in_month(valid_date) * mean)

df_seas5_adm2_seasonal <- bind_rows(
  seas5_aggregate_forecast(df_seas5_adm2_mm, value = "value_mm",
    valid_months = PRIMERA_MONTHS,
    by = c("iso3", "pcode", "issued_date")) |> mutate(window = "primera"),
  seas5_aggregate_forecast(df_seas5_adm2_mm, value = "value_mm",
    valid_months = POSTRERA_MONTHS,
    by = c("iso3", "pcode", "issued_date")) |> mutate(window = "postrera")
) |>
  rename(fcst_mm = value_mm) |>
  mutate(year = year(issued_date), issued_month = month(issued_date)) |>
  filter(
    (window == "primera" & issued_month %in% PRIMERA_ISSUED_MONTHS) |
    (window == "postrera" & issued_month %in% POSTRERA_ISSUED_MONTHS)
  )

df_era5_adm2_monthly <- df_era5_gtm_adm2 |>
  mutate(year = year(valid_date), month = month(valid_date),
         value_mm = mean * days_in_month(valid_date))

df_era5_adm2_seasonal <- bind_rows(
  df_era5_adm2_monthly |>
    filter(month %in% PRIMERA_MONTHS) |>
    group_by(pcode, iso3, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "primera"),
  df_era5_adm2_monthly |>
    filter(month %in% POSTRERA_MONTHS) |>
    group_by(pcode, iso3, year) |>
    summarise(obs_mm = sum(value_mm), .groups = "drop") |>
    mutate(window = "postrera")
)

df_adm2_joined <- df_seas5_adm2_seasonal |>
  left_join(df_era5_adm2_seasonal |> select(pcode, year, window, obs_mm),
            by = c("pcode", "year", "window")) |>
  filter(!is.na(obs_mm))

# Compute ROC-AUC per pcode/window/leadtime
drought_levels <- c("drought", "no_drought")
baseline_start <- 1981
baseline_end <- 2024
rp_target <- 4

df_adm2_baseline <- df_adm2_joined |>
  filter(year >= baseline_start, year <= baseline_end)

obs_thresh_adm2 <- df_adm2_baseline |>
  group_by(pcode, window) |>
  summarise(obs_thresh = calc_rp_threshold(obs_mm, rp_target, -1), .groups = "drop")

fcst_thresh_adm2 <- df_adm2_baseline |>
  group_by(pcode, window, leadtime) |>
  summarise(fcst_thresh = calc_rp_threshold(fcst_mm, rp_target, -1), .groups = "drop")

df_adm2_skill <- df_adm2_baseline |>
  left_join(obs_thresh_adm2, by = c("pcode", "window")) |>
  left_join(fcst_thresh_adm2, by = c("pcode", "window", "leadtime")) |>
  mutate(
    truth = fct(if_else(obs_mm <= obs_thresh, "drought", "no_drought"),
                levels = drought_levels),
    estimate = fct(if_else(fcst_mm <= fcst_thresh, "drought", "no_drought"),
                   levels = drought_levels)
  ) |>
  filter(!is.na(truth), !is.na(estimate)) |>
  group_by(pcode, window, leadtime) |>
  summarise(
    roc_auc = tryCatch(roc_auc_vec(truth, -fcst_mm, event_level = "first"),
                       error = function(e) NA_real_),
    .groups = "drop"
  )
```

```{r}
#| label: fig-gtm-adm2-roc-map
#| fig-width: 10
#| fig-height: 14
#| fig-cap: "ROC-AUC for predicting 1-in-4 year drought at admin 2 level across Guatemala (1981-2024)."

# Normalise leadtime per window
df_adm2_skill <- df_adm2_skill |>
  group_by(window) |>
  mutate(leadtime = leadtime - min(leadtime)) |>
  ungroup()

pcode_col_adm2 <- intersect(c("ADM2_PCODE", "adm2_pcode", "pcode"), names(gdf_gtm_adm2))[1]

gdf_adm2_skill <- gdf_gtm_adm2 |>
  rename(pcode = !!pcode_col_adm2) |>
  left_join(df_adm2_skill, by = "pcode") |>
  filter(!is.na(window)) |>
  mutate(
    window_label = factor(window,
      levels = c("primera", "postrera"),
      labels = c("Primera (May-Aug)", "Postrera (Sep-Nov)")
    ),
    lt_label = paste0("LT", leadtime)
  )

# Admin 1 borders for reference
gdf_adm1_border <- gdf_gtm |>
  rename(pcode = !!pcode_col) |>
  st_union() |>
  st_cast("MULTILINESTRING")

ggplot(gdf_adm2_skill) +
  geom_sf(aes(fill = roc_auc), color = "white", linewidth = 0.1) +
  geom_sf(data = gdf_gtm |> rename(pcode = !!pcode_col),
          fill = NA, color = "grey40", linewidth = 0.3) +
  facet_grid(lt_label ~ window_label) +
  scale_fill_gradient2(
    low = "#D73027", mid = "#FFFFBF", high = "#1A9850",
    midpoint = 0.65, limits = c(0.3, 1), name = "ROC-AUC",
    oob = scales::squish, na.value = "grey90"
  ) +
  labs(
    title = "Guatemala Admin 2: SEAS5 Forecast Skill (ROC-AUC)",
    subtitle = "RP4 drought | 1981-2024 baseline | Admin 1 borders in grey"
  ) +
  theme_void() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "right",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  )
```

```{r}
#| label: fig-gtm-adm2-roc-map-zoom
#| fig-width: 10
#| fig-height: 14
#| fig-cap: "ROC-AUC zoomed to Baja Verapaz & Quiché. Purple borders highlight admin 2 zones with ROC-AUC >= 0.7."

# Filter to admin 2s within Baja Verapaz (GT15) and Quiché (GT14)
zoom_adm1 <- c("GT14", "GT15")
gdf_zoom <- gdf_adm2_skill |>
  filter(substr(pcode, 1, 4) %in% zoom_adm1)

# Subset with high skill for purple highlight
gdf_high_skill <- gdf_zoom |> filter(roc_auc >= 0.7)

# Admin 1 borders within zoom area
gdf_adm1_zoom <- gdf_gtm |>
  rename(pcode = !!pcode_col) |>
  filter(pcode %in% zoom_adm1)

# Centroid labels with ROC-AUC score
gdf_zoom_centroids <- st_centroid(gdf_zoom) |>
  mutate(roc_label = sprintf("%.2f", roc_auc))

ggplot(gdf_zoom) +
  geom_sf(aes(fill = roc_auc), color = "white", linewidth = 0.1) +
  geom_sf(data = gdf_high_skill, fill = NA, color = "#9B30FF", linewidth = 0.6) +
  geom_sf(data = gdf_adm1_zoom, fill = NA, color = "black", linewidth = 0.5, alpha = 0.4) +
  geom_sf_text(data = gdf_zoom_centroids, aes(label = roc_label),
               size = 2.2, color = "grey20") +
  facet_grid(lt_label ~ window_label) +
  scale_fill_gradient2(
    low = "#D73027", mid = "#FFFFBF", high = "#1A9850",
    midpoint = 0.65, limits = c(0.3, 1), name = "ROC-AUC",
    oob = scales::squish, na.value = "grey90"
  ) +
  labs(
    title = "Baja Verapaz & Quiché: Admin 2 Forecast Skill (ROC-AUC)",
    subtitle = "RP4 drought | 1981-2024 baseline | Purple = ROC-AUC \u2265 0.7"
  ) +
  theme_void() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "right",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  )
```

```{r}
#| label: sn-load-weights
#| cache: true

# Admin 2 pixel weights for GT14+GT15
con <- pg_con()
df_weights_adm2 <- tbl(con, "polygon") |>
  mutate(across(pcode, as.character)) |>
  filter(adm_level == 2, iso3 == "GTM") |>
  select(pcode, seas5_n_upsampled_pixels) |>
  collect() |>
  filter(substr(pcode, 1, 4) %in% c("GT14", "GT15"))
DBI::dbDisconnect(con)
```

```{r}
#| label: sn-prep-data

sn_adm1 <- c("GT14", "GT15")

# Filter admin 2 joined data to GT14+GT15 (already loaded from earlier chunks)
df_sn_joined <- df_adm2_joined |>
  filter(substr(pcode, 1, 4) %in% sn_adm1) |>
  filter(year >= 1981, year <= 2024) |>
  group_by(window) |>
  mutate(leadtime = leadtime - min(leadtime)) |>
  ungroup()

sn_all_pcodes <- unique(df_sn_joined$pcode)
cat(sprintf("Admin 2 zones in GT14+GT15: %d\n", length(sn_all_pcodes)))
```

### Per-LT Leave-One-Out: Primera Offenders

For each primera lead time, we compute the change in merged ROC-AUC when each admin 2 zone is removed. Green zones are contributors (removing them hurts); red zones are offenders (removing them helps). Individual LOO deltas are small because the offending zones are spatially clustered and share correlated forecast errors — removing one barely helps when its neighbours remain.

```{r}
#| label: sn-loo-primera

# Helper: merged ROC for single window-LT
sn_merged_roc <- function(pcodes, df_joined, df_weights,
                           rp_target = 3, target_window, target_lt) {
  df_m <- df_joined |>
    filter(pcode %in% pcodes, window == target_window, leadtime == target_lt) |>
    left_join(df_weights |> filter(pcode %in% pcodes), by = "pcode") |>
    group_by(year) |>
    summarise(
      fcst_mm = weighted.mean(fcst_mm, w = seas5_n_upsampled_pixels),
      obs_mm  = weighted.mean(obs_mm,  w = seas5_n_upsampled_pixels),
      .groups = "drop"
    )
  obs_thresh <- calc_rp_threshold(df_m$obs_mm, rp_target, -1)
  df_m <- df_m |>
    mutate(truth = fct(if_else(obs_mm <= obs_thresh, "drought", "no_drought"),
                       levels = c("drought", "no_drought")))
  tryCatch(roc_auc_vec(df_m$truth, -df_m$fcst_mm, event_level = "first"),
           error = function(e) NA_real_)
}

primera_lts <- df_sn_joined |>
  filter(window == "primera") |> distinct(leadtime) |>
  arrange(leadtime) |> pull(leadtime)

# Baseline per LT
baseline_lt <- tibble(
  leadtime = primera_lts,
  baseline_roc = map_dbl(primera_lts, ~sn_merged_roc(
    sn_all_pcodes, df_sn_joined, df_weights_adm2, 3, "primera", .x))
)

# LOO per admin 2 per LT
loo_lt <- list()
for (pc in sn_all_pcodes) {
  remaining <- setdiff(sn_all_pcodes, pc)
  for (lt in primera_lts) {
    roc_without <- sn_merged_roc(remaining, df_sn_joined, df_weights_adm2, 3, "primera", lt)
    bl <- baseline_lt$baseline_roc[baseline_lt$leadtime == lt]
    loo_lt[[length(loo_lt) + 1]] <- tibble(
      pcode = pc, leadtime = lt, roc_without = roc_without,
      baseline_roc = bl, delta = roc_without - bl
    )
  }
}

df_sn_loo <- bind_rows(loo_lt) |>
  mutate(lt_label = paste0("LT", leadtime))
```

```{r}
#| label: fig-sn-loo-primera
#| fig-width: 10
#| fig-height: 24
#| fig-cap: "Leave-one-out primera skill impact per lead time (RP3 drought, 1981-2024). Green = removing zone hurts merged skill (good zone). Red = removing zone helps (offender)."

# Spatial data already loaded
pcode_col_adm2 <- intersect(c("ADM2_PCODE", "adm2_pcode", "pcode"), names(gdf_gtm_adm2))[1]
gdf_sn <- gdf_gtm_adm2 |>
  rename(pcode = !!pcode_col_adm2) |>
  filter(substr(pcode, 1, 4) %in% sn_adm1)

gdf_sn_adm1 <- gdf_gtm |>
  rename(pcode = !!pcode_col) |>
  filter(pcode %in% sn_adm1)

loo_maps <- list()
for (lt in primera_lts) {
  gdf_lt <- gdf_sn |>
    left_join(df_sn_loo |> filter(leadtime == lt) |> select(pcode, delta),
              by = "pcode")
  gdf_lt_c <- st_centroid(gdf_lt) |>
    mutate(delta_label = sprintf("%+.3f", delta))

  bl_roc <- baseline_lt$baseline_roc[baseline_lt$leadtime == lt]

  loo_maps[[paste0("LT", lt)]] <- ggplot(gdf_lt) +
    geom_sf(aes(fill = delta), color = "white", linewidth = 0.3) +
    geom_sf(data = gdf_sn_adm1, fill = NA, color = "black",
            linewidth = 0.7, alpha = 0.5) +
    geom_sf_text(data = gdf_lt_c, aes(label = delta_label),
                 size = 3.5, color = "grey20", fontface = "bold") +
    scale_fill_gradient2(
      low = "#D73027", mid = "white", high = "#1A9850",
      midpoint = 0, name = "Delta\nROC-AUC",
      limits = c(-0.04, 0.04), oob = scales::squish
    ) +
    labs(title = sprintf("Primera LT%d (baseline ROC = %.3f)", lt, bl_roc)) +
    theme_void() +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "right",
      legend.text = element_text(size = 10),
      legend.title = element_text(size = 11)
    )
}

patchwork::wrap_plots(loo_maps, ncol = 1) +
  patchwork::plot_annotation(
    title = "LOO: primera skill change when admin 2 removed (per lead time, RP3)",
    subtitle = "Green = removing hurts (good zone) | Red = removing helps (offender)",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10)
    )
  )
```

### Forecast Similarity Clustering

To address the limitation of single-zone LOO (offenders are spatially correlated, so individual deltas are tiny), we cluster admin 2 zones by forecast similarity and evaluate skill when entire clusters are removed.

We compute pairwise Spearman correlation of primera forecast time series across all admin 2 zones, then apply hierarchical clustering (Ward's D2 on the distance matrix `1 - correlation`). We evaluate cuts at k = 2 and k = 3 to find a partition that cleanly separates the low-skill block.

```{r}
#| label: sn-forecast-clustering

# Build primera forecast matrix: rows = years, cols = admin 2 zones
# Average across lead times for the clustering
df_primera_wide <- df_sn_joined |>
  filter(window == "primera") |>
  group_by(pcode, year) |>
  summarise(fcst_mm = mean(fcst_mm), .groups = "drop") |>
  pivot_wider(names_from = pcode, values_from = fcst_mm)

fcst_mat <- df_primera_wide |> select(-year) |> as.matrix()
rownames(fcst_mat) <- df_primera_wide$year

# Spearman correlation → distance → hierarchical clustering
cor_mat <- cor(fcst_mat, method = "spearman", use = "complete.obs")
dist_mat <- as.dist(1 - cor_mat)
hc <- hclust(dist_mat, method = "ward.D2")

# Get zone names column
sn_name_col <- intersect(
  c("ADM2_ES", "adm2_es", "ADM2_EN", "adm2_en", "shapeName"),
  names(gdf_sn)
)[1]

# Cut at k = 2 and k = 3
df_clusters_k2 <- tibble(pcode = names(cutree(hc, k = 2)),
                          cluster = cutree(hc, k = 2))
df_clusters_k3 <- tibble(pcode = names(cutree(hc, k = 3)),
                          cluster = cutree(hc, k = 3))
```

```{r}
#| label: fig-sn-cluster-maps
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Admin 2 zones coloured by forecast similarity cluster at k=2 (left) and k=3 (right). Hierarchical clustering (Ward's D2) on Spearman correlation of primera SEAS5 forecasts."

cluster_pal <- c(
  "1" = "#E41A1C", "2" = "#377EB8",
  "3" = "#4DAF4A", "4" = "#984EA3"
)

make_cluster_map <- function(df_cl, k_label) {
  gdf_cl <- gdf_sn |>
    left_join(df_cl, by = "pcode") |>
    mutate(cluster_label = factor(cluster))

  gdf_cl_c <- st_centroid(gdf_cl) |>
    rename(zone_name = !!sn_name_col)

  ggplot(gdf_cl) +
    geom_sf(aes(fill = cluster_label), color = "white", linewidth = 0.3) +
    geom_sf(data = gdf_sn_adm1, fill = NA, color = "black",
            linewidth = 0.7, alpha = 0.5) +
    geom_sf_text(data = gdf_cl_c, aes(label = zone_name),
                 size = 2.2, color = "grey20") +
    scale_fill_manual(values = cluster_pal, name = "Cluster", drop = FALSE) +
    labs(title = k_label) +
    theme_void() +
    theme(
      plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
      legend.position = "right"
    )
}

p_k2 <- make_cluster_map(df_clusters_k2, "k = 2")
p_k3 <- make_cluster_map(df_clusters_k3, "k = 3")

patchwork::wrap_plots(p_k2, p_k3, ncol = 2) +
  patchwork::plot_annotation(
    title = "Forecast Similarity Clusters (Primera)",
    subtitle = "Ward's D2 on Spearman correlation of SEAS5 primera forecasts",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10)
    )
  )
```

```{r}
#| label: sn-leave-cluster-out

# For each k and each cluster, compute merged primera skill without it
build_cluster_configs <- function(df_cl, k_label) {
  configs <- list()

  # Individual cluster removal
  for (cl in sort(unique(df_cl$cluster))) {
    cl_zones <- df_cl |> filter(cluster == cl) |> pull(pcode)
    remaining <- setdiff(sn_all_pcodes, cl_zones)
    n_removed <- length(cl_zones)
    configs[[sprintf("%s: Drop CL%d (%d zones)", k_label, cl, n_removed)]] <- remaining
  }

  # Multi-cluster removal: all 2-cluster combos
  all_cls <- sort(unique(df_cl$cluster))
  if (length(all_cls) >= 2) {
    for (i in seq_along(all_cls)) {
      for (j in seq_along(all_cls)) {
        if (j <= i) next
        cl_i <- all_cls[i]; cl_j <- all_cls[j]
        cl_zones <- df_cl |> filter(cluster %in% c(cl_i, cl_j)) |> pull(pcode)
        remaining <- setdiff(sn_all_pcodes, cl_zones)
        if (length(remaining) < 4) next
        n_removed <- length(cl_zones)
        configs[[sprintf("%s: Drop CL%d+CL%d (%d zones)", k_label,
                         cl_i, cl_j, n_removed)]] <- remaining
      }
    }
  }

  configs
}

sn_cluster_configs <- c(
  list("All 29 zones (baseline)" = sn_all_pcodes),
  build_cluster_configs(df_clusters_k2, "k=2"),
  build_cluster_configs(df_clusters_k3, "k=3")
)

# Compute skill for each config
sn_cluster_skill <- list()
for (nm in names(sn_cluster_configs)) {
  pcs <- sn_cluster_configs[[nm]]
  for (lt in primera_lts) {
    roc_val <- sn_merged_roc(pcs, df_sn_joined, df_weights_adm2, 3, "primera", lt)
    sn_cluster_skill[[length(sn_cluster_skill) + 1]] <- tibble(
      config = nm, leadtime = lt, roc_auc = roc_val,
      n_zones = length(pcs)
    )
  }
}

df_cluster_skill <- bind_rows(sn_cluster_skill)
```

```{r}
#| label: tbl-sn-cluster-out
#| tbl-cap: "Leave-cluster-out analysis: primera ROC-AUC (RP3) when clusters are removed. Shown for k=2 and k=3 cuts of the dendrogram."

# Build cluster membership footnotes for both k values
make_cluster_footnote <- function(df_cl, k_label) {
  df_cl |>
    left_join(
      gdf_sn |> st_drop_geometry() |>
        select(pcode, any_of(sn_name_col)) |>
        rename(name = !!sn_name_col),
      by = "pcode"
    ) |>
    group_by(cluster) |>
    summarise(zones = paste(name, collapse = ", "), n = n(), .groups = "drop") |>
    mutate(label = sprintf("%s CL%d (%d): %s", k_label, cluster, n, zones)) |>
    pull(label) |>
    paste(collapse = ". ")
}

footnote_k2 <- make_cluster_footnote(df_clusters_k2, "k=2")
footnote_k3 <- make_cluster_footnote(df_clusters_k3, "k=3")

df_cluster_skill |>
  mutate(
    lt_label = paste0("LT", leadtime),
    roc_label = sprintf("%.3f", roc_auc)
  ) |>
  select(config, n_zones, lt_label, roc_label) |>
  pivot_wider(names_from = lt_label, values_from = roc_label) |>
  arrange(desc(n_zones)) |>
  gt() |>
  tab_header(
    title = md("**Leave-Cluster-Out: Primera Skill**"),
    subtitle = "ROC-AUC | RP3 drought | 1981-2024 baseline"
  ) |>
  cols_label(config = "Configuration", n_zones = "Zones") |>
  tab_footnote(footnote = footnote_k2) |>
  tab_footnote(footnote = footnote_k3)
```

The clustering confirms that the low-skill zones concentrate in a geographically coherent block in northern Quiché. However, no single cluster removal fully captures the set of offenders — the low-skill zones span multiple forecast-similarity clusters, and the spatial correlation of forecast errors makes automated selection difficult. This supports the simpler approach adopted in the main recommendation: use the merged AOI where skill is adequate (primera LT0-LT1, all postrera), and fall back to Baja Verapaz only for primera LT2.

### Qualitative Admin 2 Removal

As an alternative to automated clustering, we manually identified 7 low-skill Quiché admin 2 zones informed by the LOO and greedy analyses above. Removing these zones yields a 22-zone AOI. The table below compares ROC-AUC across all lead times and seasons for three configurations: the full 29-zone merge, the 22-zone qualitative selection, and Baja Verapaz (GT15) only.

```{r}
#| label: sn-qual-removal

qual_remove <- c("GT1413", "GT1411", "GT1410", "GT1416", "GT1409", "GT1415", "GT1405")
qual_remaining <- setdiff(sn_all_pcodes, qual_remove)
gt15_only <- sn_all_pcodes[startsWith(sn_all_pcodes, "GT15")]

postrera_lts <- df_sn_joined |>
  filter(window == "postrera") |> distinct(leadtime) |>
  arrange(leadtime) |> pull(leadtime)

aoi_configs <- list(
  "All 29 zones (GT14+GT15)" = sn_all_pcodes,
  "22 zones (7 Quiché removed)" = qual_remaining,
  "GT15 only (Baja Verapaz)" = gt15_only
)

qual_results <- list()
for (rp in c(3, 4)) {
  for (nm in names(aoi_configs)) {
    pcs <- aoi_configs[[nm]]
    for (lt in primera_lts) {
      roc_val <- sn_merged_roc(pcs, df_sn_joined, df_weights_adm2, rp, "primera", lt)
      qual_results[[length(qual_results) + 1]] <- tibble(
        config = nm, rp_def = rp, window = "primera", leadtime = lt,
        roc_auc = roc_val, n_zones = length(pcs)
      )
    }
    for (lt in postrera_lts) {
      roc_val <- sn_merged_roc(pcs, df_sn_joined, df_weights_adm2, rp, "postrera", lt)
      qual_results[[length(qual_results) + 1]] <- tibble(
        config = nm, rp_def = rp, window = "postrera", leadtime = lt,
        roc_auc = roc_val, n_zones = length(pcs)
      )
    }
  }
}

df_qual <- bind_rows(qual_results)
```

```{r}
#| label: fig-sn-qual-removal-map
#| fig-cap: "Qualitative admin 2 selection for the 22-zone AOI. Green = retained (22 zones), red = removed (7 low-skill Quiché zones)."
#| fig-width: 8
#| fig-height: 8

gdf_qual <- gdf_sn |>
  mutate(
    status = if_else(pcode %in% qual_remove, "Removed", "Retained"),
    dept = if_else(startsWith(pcode, "GT15"), "Baja Verapaz", "Quiché")
  )

gdf_qual_c <- st_centroid(gdf_qual) |>
  rename(zone_name = !!sn_name_col)

ggplot(gdf_qual) +
  geom_sf(aes(fill = status), color = "white", linewidth = 0.3) +
  geom_sf(data = gdf_sn_adm1, fill = NA, color = "black",
          linewidth = 0.7, alpha = 0.5) +
  geom_sf_text(data = gdf_qual_c, aes(label = zone_name),
               size = 2.2, color = "grey20") +
  scale_fill_manual(
    values = c("Retained" = "#1A9850", "Removed" = "#D73027"),
    name = "Status"
  ) +
  labs(
    title = "Qualitative Admin 2 Selection (GT14 + GT15)",
    subtitle = "7 low-skill Quiché zones removed → 22-zone AOI"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    legend.position = "right"
  )
```

```{r}
#| label: tbl-sn-qual-removal
#| tbl-cap: "ROC-AUC for three AOI configurations across all lead times and seasons at RP3 and RP4 drought definitions."

# Get zone names for removed zones
qual_names <- gdf_sn |>
  st_drop_geometry() |>
  filter(pcode %in% qual_remove) |>
  select(pcode, any_of(sn_name_col))

qual_footnote <- sprintf("Removed zones: %s",
  paste(sprintf("%s (%s)", qual_names[[sn_name_col]], qual_names$pcode), collapse = ", "))

make_qual_gt <- function(df, rp_val) {
  df |>
    filter(rp_def == rp_val) |>
    mutate(
      lt_label = sprintf("%s LT%d", tools::toTitleCase(window), leadtime),
      roc_label = sprintf("%.3f", roc_auc)
    ) |>
    select(config, n_zones, lt_label, roc_label) |>
    pivot_wider(names_from = lt_label, values_from = roc_label) |>
    gt() |>
    tab_header(
      title = md(sprintf("**Qualitative Removal: Skill Across Seasons & Lead Times (RP%d)**", rp_val)),
      subtitle = sprintf("ROC-AUC | RP%d drought definition", rp_val)
    ) |>
    cols_label(config = "Configuration", n_zones = "Zones") |>
    tab_spanner(label = "Primera", columns = starts_with("Primera")) |>
    tab_spanner(label = "Postrera", columns = starts_with("Postrera")) |>
    tab_footnote(footnote = qual_footnote)
}

make_qual_gt(df_qual, 4)
```

```{r}
#| label: tbl-sn-qual-removal-rp3

make_qual_gt(df_qual, 3)
```
