# 2026 Situation Summary {#sec-rolac-summary}

```{r}
#| label: setup-summary
#| include: false
library(tidyverse); library(sf); library(patchwork); library(gghdx); library(scales); library(cumulus)
gghdx(); sf::sf_use_s2(FALSE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Derived inputs are read from blob (projects / ds-aa-lac-dry-corridor/book/static/). The scripts
# that produce them are tracked in data-raw/mission_support_2026/. Re-rendering needs the read SAS
# (DSCI_AZ_BLOB_DEV_SAS); the committed _freeze/ output lets the book be viewed without credentials.
rd <- function(f) cumulus::blob_read(paste0("ds-aa-lac-dry-corridor/book/static/", f),
                                     stage = "dev", container = "projects", progress_show = FALSE)
obs    <- rd("primera_observed_2026.csv")        # rp_dry
mixed  <- rd("mixed_primera_2026.csv")           # z_obs, z_mixed, rp_emp
joint  <- rd("primera_vs_postrera_2026.csv")     # z_post, rp_post
asi    <- rd("asi_primera_2026.csv")             # asi_mean_2026, asi_pctile
season <- rd("season_specific_risk_adm1.csv")    # asi_2026, psp, crop/psp gates, season_class
asi_ef <- rd("early_vs_final_asi.csv")           # early vs final ASI
pnorm  <- rd("pct_normal_2026.csv")              # % of normal
xwalk  <- rd("priority_synthesis_adm1.csv") |> select(pcode, gaul_code)
gaul   <- rd("gaul_adm1_gtm_hnd_slv.geojson") |> janitor::clean_names() |>
  sf::st_make_valid() |> (\(g) sf::st_collection_extract(g, "POLYGON"))()

iso_lab <- c(GTM = "Guatemala", HND = "Honduras", SLV = "El Salvador")
join_gaul <- function(df, by = "gaul_code") left_join(gaul, df, by = c("adm1_code" = by)) |>
  mutate(iso3c = c(Guatemala = "GTM", Honduras = "HND", `El Salvador` = "SLV")[adm0_name])

# --- palettes --------------------------------------------------------------
# ASI (stress): green only at the very bottom, then reds dominate (green->yellow->orange->red->dark).
# Use anchored breakpoints so moderate stress already reads red, per the "mostly red" intent.
asi_pal <- c("#1a9850", "#fee08b", "#fdae61", "#e34a33", "#7f0000")
asi_scale <- function(anchors, name) {
  lim <- range(anchors)
  scale_fill_gradientn(colours = asi_pal, values = scales::rescale(anchors, from = lim),
                       limits = lim, na.value = "grey70", name = name)
}
# precipitation ANOMALY (z): red (drier/below) -> white -> blue (wetter/above)
precip_scale <- function(z, name = "Rainfall anomaly (σ)") {
  m <- max(abs(z), na.rm = TRUE)
  scale_fill_gradient2(low = "#b2182b", mid = "#f7f7f7", high = "#2166ac", midpoint = 0,
                       limits = c(-m, m), na.value = "grey85", name = name)
}

# 3 countries stacked (1 col x 3 rows -> big), each fit to its own extent; shared bottom legend.
# outline_sf (optional): sf carrying iso3c (+ `oflag` if outline_map given) — drawn PER COUNTRY so
# outlines never bleed across borders. outline_map = named colours for a categorical outline legend;
# NULL -> a single fixed black outline.
land3 <- function(df, layers, title = NULL, outline_sf = NULL, outline_map = NULL) {
  df <- filter(df, !is.na(iso3c))
  mk <- function(cc) {
    cd <- filter(df, iso3c == cc); b <- st_bbox(cd)
    p <- ggplot(cd) + layers
    if (!is.null(outline_sf)) {
      oc <- filter(outline_sf, iso3c == cc)
      if (nrow(oc) > 0) p <- p + (if (is.null(outline_map))
        geom_sf(data = oc, fill = NA, color = "black", linewidth = 0.45)
      else geom_sf(data = oc, aes(color = oflag), fill = NA, linewidth = 0.7))
    }
    if (!is.null(outline_map)) p <- p +
      scale_color_manual(values = outline_map, drop = FALSE, limits = names(outline_map), name = NULL)
    p + coord_sf(xlim = c(b["xmin"], b["xmax"]), ylim = c(b["ymin"], b["ymax"])) +
      labs(subtitle = iso_lab[cc]) +
      theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank(),
            plot.subtitle = element_text(face = "bold", size = 13), legend.position = "bottom")
  }
  wrap_plots(lapply(c("GTM","HND","SLV"), mk), ncol = 1, guides = "collect") +
    plot_annotation(title = title, theme = theme(plot.title = element_text(face = "bold", size = 14)))
}

# Two metrics side by side (left | right), 3 countries stacked. Country labels on the left column.
pair3 <- function(df, left_layers, right_layers, title = NULL, right_outline = NULL) {
  df <- filter(df, !is.na(iso3c))
  cell <- function(cc, layers, show_country, outline = NULL) {
    cd <- filter(df, iso3c == cc); b <- st_bbox(cd)
    p <- ggplot(cd) + layers
    if (!is.null(outline)) {
      oc <- filter(outline, iso3c == cc)
      if (nrow(oc) > 0) p <- p + geom_sf(data = oc, fill = NA, color = "black", linewidth = 0.45)
    }
    p + coord_sf(xlim = c(b["xmin"], b["xmax"]), ylim = c(b["ymin"], b["ymax"])) +
      labs(subtitle = if (show_country) iso_lab[cc] else NULL) +
      theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank(),
            plot.subtitle = element_text(face = "bold", size = 12), legend.position = "bottom")
  }
  plots <- unlist(lapply(c("GTM","HND","SLV"),
    function(cc) list(cell(cc, left_layers, TRUE), cell(cc, right_layers, FALSE, right_outline))), recursive = FALSE)
  wrap_plots(plots, ncol = 2, byrow = TRUE, guides = "collect") +
    plot_annotation(title = title, theme = theme(plot.title = element_text(face = "bold", size = 14)))
}
```

## Purpose {.unnumbered}

This chapter summarizes CHD's desk review for the July 2026 ROLAC missions to El Salvador,
Honduras, and Guatemala. It sets out the evidence behind our recommendations on where and how to
act. Supporting detail is in the companion chapter.

::: {.callout-note}
## Bottom line
This analysis characterizes the drought and its agricultural impact to inform in-country,
cross-agency planning. It is not intended to set operational priorities on its own.

Anticipatory Action for the 2026 primera was activated in March 2026 on the seasonal forecast. The
severe, regional drought now developing indicates that the forecast identified the signal correctly
and that the framework acted ahead of the impact now being observed.

1. The drought is regional and largely undifferentiated. Rainfall is well below normal across
   nearly all admin-1 units in the three countries.
2. Because the hazard is largely undifferentiated, the drought signal alone in this analysis is not
   a sound basis for ranking where to act. Prioritization is better anchored in operational capacity
   and local vulnerability, assembled by pooling in-country, cross-agency expertise.
3. Crops are already stressed and the season is unlikely to recover. The postrera harvest is also
   likely to be affected, though that outlook is less certain.
4. Seasonal timing constrains the options available. The primera is already underway, so the window
   for additional agricultural-based anticipatory action there has closed and mitigation is what
   remains, whereas the postrera retains lead time for anticipatory measures. Areas whose crops fall outside the primera and
   postrera calendar are better considered through a separate and additional vulnerability lens than
   the rest. These distinctions are offered to inform in-country decisions, not to prescribe them.
:::

# Current situation

## Observed rainfall

Observed May–June rainfall (ERA5) was well below normal in every admin-1 unit of all three
countries, with the largest deficits in Honduras and eastern Guatemala. The dryness is broad rather
than confined to isolated pockets.

```{r}
#| label: fig-obs-anom
#| fig-cap: "Observed May–June 2026 rainfall anomaly (ERA5, standard deviations from normal). Red = drier than normal, blue = wetter."
#| fig-width: 7.5
#| fig-height: 11
d <- mixed |> left_join(xwalk, by = "pcode") |> join_gaul()
land3(d, list(geom_sf(aes(fill = z_obs), color = "grey80", linewidth = 0.08), precip_scale(d$z_obs)),
      "Observed May–June rainfall anomaly")
```

## Agricultural stress

The FAO Agricultural Stress Index (ASI), the share of a unit's crop area in severe stress, shows
that impact is already occurring. Stress is concentrated in Honduras (Olancho 52%, El Paraíso 37%,
Choluteca 25%) and eastern El Salvador (San Vicente 22%, La Unión 18%, Morazán 16%). Guatemala
remains mostly low, with Zacapa (16%) the main exception.

The same stress can also be expressed as a return period, measured against each unit's own
1984–2025 record. On that basis the current stress is rare in the most affected units, on the order
of one in fifteen years in Olancho and San Vicente. The magnitude and return-period maps are shown
side by side below.

```{r}
#| label: fig-asi-pair
#| fig-cap: "Agricultural Stress Index, late June 2026. Left: magnitude — % of crop area in severe stress (green = low, red = high). Right: rarity — return period of 2026's stress vs each unit's own 1984–2025 record, Weibull (green = ordinary, red = rare); black outline = units whose 2026 stress is in the worst (lower) tercile of their own record. Grey = no ASI / insufficient cropland."
#| fig-width: 10
#| fig-height: 12
a <- asi |> transmute(gaul_code = adm1_code, asi = asi_mean_2026, asi_pctile, asi_rp = 1 / (1 - asi_pctile)) |> join_gaul()
amax <- max(a$asi, na.rm = TRUE)
rpmax <- min(max(a$asi_rp, na.rm = TRUE), 30)
worst_tercile <- filter(a, asi_pctile >= 2/3)   # 2026 in the unit's own worst third
pair3(a,
      left_layers  = list(geom_sf(aes(fill = asi), color = "grey80", linewidth = 0.08),
                          asi_scale(c(0, 3, 8, 20, amax), "ASI (% crop area stressed)")),
      right_layers = list(geom_sf(aes(fill = pmin(asi_rp, 30)), color = "grey80", linewidth = 0.08),
                          asi_scale(c(1, 1.5, 2.5, 8, rpmax), "ASI return period (yr)")),
      title = "ASI — magnitude (left) and return period (right)",
      right_outline = worst_tercile)
```

# Primera recovery unlikely

## Complete-season outlook (observed + forecast)

Combining the observed May–June rainfall with the July–August forecast gives the complete primera.
On that basis the season remains a rare deficit across the region, and no unit shows a wet signal
that would offset the current shortfall.

```{r}
#| label: fig-mixed-pair
#| fig-cap: "Combined observed (May–June) and forecast (July–August) primera. Left: rainfall as percent-anomaly (σ from normal; red drier, blue wetter). Right: the corresponding empirical return period (pale to dark red = rarer)."
#| fig-width: 10
#| fig-height: 12
d <- mixed |> left_join(xwalk, by = "pcode") |> mutate(rp_c = pmin(rp_emp, 30)) |> join_gaul()
pair3(d,
  left_layers  = list(geom_sf(aes(fill = z_mixed), color = "grey80", linewidth = 0.08),
                      precip_scale(d$z_mixed, "Anomaly (σ)")),
  right_layers = list(geom_sf(aes(fill = rp_c), color = "grey80", linewidth = 0.08),
                      scale_fill_gradient(low = "#fff5f0", high = "#67000d", limits = c(1, 30),
                                          na.value = "grey90", name = "Return period (yr)")))
```

## Early stress as a predictor

End-June stress is a leading indicator, not only a current reading. Across the 1984–2025 record,
end-June ASI is correlated with the eventual season (r = 0.75), and a unit in its own worst tercile
at end-June goes on to a worst-tercile season about 72% of the time, against a 33% base rate. Where
crops are stressed now, the primera is therefore likely to fail.

```{r}
#| label: fig-predict
#| fig-cap: "End-June ASI vs the eventual Jul–Oct peak, each admin-1 × year 1984–2025. Strong positive relationship (r = 0.75)."
#| fig-width: 8
#| fig-height: 5
hist <- asi_ef |> filter(year < 2026)
ggplot(hist, aes(early, final)) +
  geom_point(alpha = 0.35, size = 1.2, color = "grey40") +
  geom_smooth(method = "lm", se = TRUE, color = "#d73027", linewidth = 0.8) +
  labs(title = "End-June stress predicts the rest of the season (r = 0.75)",
       x = "End-June ASI (early signal, % crop area stressed)", y = "Peak ASI Jul–Oct (season outcome)")
```

## Postrera also looks unfavourable {#sec-postrera-outlook}

The September–November (postrera) forecast is also poor across the region, so the second harvest is
unlikely to compensate for primera losses. Forecast skill for the postrera is weak, so this is best
read as an outlook rather than a firm prediction.

```{r}
#| label: fig-postrera-pair
#| fig-cap: "SEAS5 postrera (Sep–Nov) forecast. Left: rainfall as percent-anomaly (σ from normal; red drier, blue wetter). Right: the corresponding empirical return period (pale to dark red = rarer). Forecast skill is weak, so this is an outlook."
#| fig-width: 10
#| fig-height: 12
dp <- joint |> left_join(xwalk, by = "pcode") |> mutate(rp_c = pmin(rp_post, 30)) |> join_gaul()
pair3(dp,
  left_layers  = list(geom_sf(aes(fill = z_post), color = "grey80", linewidth = 0.08),
                      precip_scale(dp$z_post, "Anomaly (σ)")),
  right_layers = list(geom_sf(aes(fill = rp_c), color = "grey80", linewidth = 0.08),
                      scale_fill_gradient(low = "#fff5f0", high = "#67000d", limits = c(1, 30),
                                          na.value = "grey90", name = "Return period (yr)")))
```

## All indicators by admin-1

The table below compiles the indicators for each admin-1 unit, grouped by country. Rainfall is
given as percent of the 1991–2024 normal alongside its empirical return period, and agricultural
stress as ASI alongside its return period. Cell colours follow the maps: rainfall from red (drier)
through white to blue (wetter), return periods from pale to dark red, and ASI from green to red.

```{r}
#| label: tbl-all
#| tbl-cap: "2026 primera and postrera indicators by admin-1. % norm = percent of the 1991–2024 normal (below 100 = drier); RP = empirical return period in years; ASI = % of crop area in severe stress."
library(gt)
tab <- mixed |> transmute(pcode, name, iso3, comb_rp = rp_emp) |>
  left_join(pnorm, by = "pcode") |>
  left_join(obs   |> select(pcode, obs_rp = rp_dry), by = "pcode") |>
  left_join(joint |> select(pcode, post_rp = rp_post), by = "pcode") |>
  left_join(xwalk, by = "pcode") |>
  left_join(asi |> transmute(gaul_code = adm1_code, asi_pct = asi_mean_2026,
                             asi_rp = 1 / (1 - asi_pctile)), by = "gaul_code") |>
  left_join(season |> transmute(gaul_code, off_cal = season_class == "Off-calendar"), by = "gaul_code") |>
  mutate(country = iso_lab[iso3],
         off_cal = tidyr::replace_na(off_cal, FALSE),
         name = ifelse(off_cal, paste0(name, " †"), name)) |>
  arrange(country, desc(asi_pct)) |>
  select(country, name, obs_pnorm, obs_rp, comb_pnorm, comb_rp, asi_pct, asi_rp, post_pnorm, post_rp, off_cal)

pdev   <- max(abs(c(tab$obs_pnorm, tab$comb_pnorm, tab$post_pnorm) - 100), na.rm = TRUE)
rpmaxr <- max(c(tab$obs_rp, tab$comb_rp, tab$post_rp), na.rm = TRUE)
pnorm_fn <- scales::col_numeric(c("#b2182b", "#f7f7f7", "#2166ac"), domain = c(100 - pdev, 100 + pdev), na.color = "grey90")
rp_fn    <- scales::col_numeric(c("#fff5f0", "#d73027", "#67000d"), domain = c(1, rpmaxr), na.color = "grey90")
asi_fn   <- scales::col_numeric(asi_pal, domain = c(0, max(tab$asi_pct, na.rm = TRUE)), na.color = "grey90")
asirp_fn <- scales::col_numeric(asi_pal, domain = c(1, max(tab$asi_rp, na.rm = TRUE)), na.color = "grey90")

gt(tab, groupname_col = "country", rowname_col = "name") |>
  tab_spanner("Primera", columns = c(obs_pnorm, obs_rp, comb_pnorm, comb_rp, asi_pct, asi_rp)) |>
  tab_spanner("Postrera", columns = c(post_pnorm, post_rp)) |>
  cols_label(obs_pnorm = "Obs rain (% norm)", obs_rp = "Obs RP", comb_pnorm = "Outlook (% norm)",
             comb_rp = "Outlook RP", asi_pct = "ASI %", asi_rp = "ASI RP",
             post_pnorm = "Fcst (% norm)", post_rp = "Fcst RP") |>
  fmt_percent(c(obs_pnorm, comb_pnorm, post_pnorm, asi_pct), decimals = 0, scale_values = FALSE) |>
  fmt_number(c(obs_rp, comb_rp, asi_rp, post_rp), decimals = 0) |>
  data_color(columns = c(obs_pnorm, comb_pnorm, post_pnorm), fn = pnorm_fn) |>
  data_color(columns = c(obs_rp, comb_rp, post_rp), fn = rp_fn) |>
  data_color(columns = asi_pct, fn = asi_fn) |>
  data_color(columns = asi_rp, fn = asirp_fn) |>
  cols_hide(off_cal) |>
  tab_style(style = cell_text(color = "grey55", style = "italic"),
            locations = list(cells_body(rows = off_cal), cells_stub(rows = off_cal))) |>
  tab_source_note(gt::md(
    "**†  Off-calendar** — the drought window falls outside the crop's rain-critical period; the seasonal-drought lens does not apply, so these rows are set aside (read with caution). Blank ASI = no cropland (Islas de la Bahía).")) |>
  sub_missing(missing_text = "—") |>
  tab_options(table.font.size = "11px", data_row.padding = "2px", row_group.font.weight = "bold")
```

# Response by season

The distinctions below follow from the seasonal timing and are meant to inform in-country planning
rather than to prescribe it.

## Primera

Because the primera is already underway, the window for additional agricultural-based anticipatory
action has closed and mitigation is the response that remains available. The map below shows current stress (ASI) and
outlines two groups: units already very stressed, where continued decline is likely, and units not
yet stressed but likely to decline given the dry outlook. Grey units have no ASI or fall off the
seasonal calendar.

```{r}
#| label: fig-primera-action
#| fig-cap: "Primera situation and outlook. Fill = ASI (green to red). Red outline = already very stressed (ASI ≥ 15%, continued decline likely); blue outline = not yet stressed but likely to decline. Grey = off-calendar or non-crop units, set aside because the seasonal-drought lens does not apply there (they are not predicted to be spared)."
#| fig-width: 7.5
#| fig-height: 11
pr <- asi |> transmute(gaul_code = adm1_code, asi = asi_mean_2026) |>
  full_join(season |> select(gaul_code, asi_2026, psp_relevant, crop_exposed), by = "gaul_code") |>
  mutate(inseason = psp_relevant %in% TRUE & crop_exposed %in% TRUE,
         asi = ifelse(inseason, asi, NA_real_),   # grey out off-calendar / non-crop: set aside, not "safe"
         flag = case_when(inseason & asi_2026 >= 15 ~ "Already very stressed — decline likely",
                          inseason & asi_2026 <  15 ~ "Not yet stressed — likely to decline",
                          TRUE ~ NA_character_)) |> join_gaul()
flagged <- filter(pr, !is.na(flag)) |> rename(oflag = flag)
out_pal <- c("Already very stressed — decline likely" = "#d7191c",
             "Not yet stressed — likely to decline" = "#2166ac")
amax <- max(pr$asi, na.rm = TRUE)
land3(pr, list(
  geom_sf(aes(fill = asi), color = "grey80", linewidth = 0.06),
  asi_scale(c(0, 3, 8, 20, amax), "ASI (% crop area)")),
  "Primera stress and expected decline",
  outline_sf = flagged, outline_map = out_pal)
```

## Postrera

Because the postrera has not yet started, lead time remains for anticipatory measures, independent
of whether the AA framework was triggered earlier this season. Forecast skill is weak, so this
supports acting ahead of an elevated but uncertain risk rather than a firm prediction of failure.
The postrera outlook is described in @sec-postrera-outlook and mapped in @fig-postrera-pair.

## Areas off the seasonal calendar

A small number of units fall outside the drought's rain-critical calendar. In the western highlands
and along the Caribbean fringe, the crop's precipitation-sensitive period (ASAP PSP) does not
coincide with the primera and postrera windows. A seasonal-rainfall deficit is not the relevant
measure of risk in these areas; they are better addressed through vulnerability and food-security
assessment than through agricultural drought response. That layer has not yet been built here.

```{r}
#| label: fig-psp
#| fig-cap: "Overlap of the drought windows with the crop's rain-critical period (ASAP PSP). Green = drought hits the rain-sensitive months (seasonal lens applies); pale = off-calendar (highlands / Caribbean)."
#| fig-width: 7.5
#| fig-height: 11
p <- season |> transmute(gaul_code, psp) |> join_gaul()
land3(p, list(geom_sf(aes(fill = psp), color = "grey80", linewidth = 0.08),
              scale_fill_gradient(low = "grey88", high = "#1a9850", na.value = "grey85",
                                  limits = c(0, 1), labels = percent, name = "Drought in rain-critical period")),
      "Where the seasonal-drought lens applies")
```

# External validation: FAO/JRC ASAP

The [Anomaly Hot Spots of Agricultural Production (ASAP)](https://agricultural-production-hotspots.ec.europa.eu/wexplorer/)
system, run by FAO and the EC Joint Research Centre, provides an independent check on the assessment
above. ASAP combines vegetation health,
biomass, and a crop water-balance model over cropland, and issues monitoring and forecast warnings
by agro-climatic zone. Its warning for the third dekad of June 2026, the same window as this
analysis, is shown below. Because it comes from a different institution and a different method,
agreement carries weight.

![FAO/JRC ASAP agricultural warning, third dekad of June 2026. Red indicates a combined water-balance and biomass warning and darker red an end-of-season biomass warning; yellow and orange indicate a water-balance or biomass warning alone. Hatching marks a forecast warning for the current and/or upcoming season. White indicates insufficient crop area.](asap_cadc.png)

ASAP corroborates the analysis on each of its main points.

- **Regional severity.** The most severe class, a combined water-balance and biomass warning, covers
  the dry corridor almost continuously across the three countries. This matches the region-wide
  rainfall deficit and the concentration of agricultural stress reported here.
- **Hotspot geography.** ASAP's strongest warnings fall in central and southern Honduras, the
  eastern-Guatemala Motagua valley, and El Salvador, the same areas identified by ASI (Olancho, El
  Paraíso, Choluteca, Zacapa, and eastern El Salvador).
- **Current and upcoming season.** ASAP issues forecast warnings for both the current and the
  upcoming season across much of the region. This independently supports the two conclusions drawn
  here: that the primera is unlikely to recover, and that the postrera outlook is also unfavourable.
  The latter is the point that rests on a lower-skill forecast in our own analysis, so the
  independent agreement is useful.
- **Cropland extent.** ASAP marks the Caribbean northeast, including Gracias a Dios, as insufficient
  crop area, consistent with the units set aside here as off-calendar or non-agricultural.

The two products use different spatial units and different warning categories, so the comparison is
close rather than unit-for-unit. Within those limits, an independent operational system reaches the
same regional assessment.

# Definitions and caveats {.unnumbered}

- **ASI**: the percentage of a unit's crop area in severe drought stress (FAO; absolute and
  comparable across units). "Already very stressed" refers to ASI of 15% or more.
- **Percent of normal**: seasonal rainfall as a share of the 1991–2024 mean; values below 100 are
  drier than normal.
- **Return period**: the empirical recurrence interval of the 2026 value against the unit's own
  record (Weibull), in years.
- **Baselines**: rainfall anomalies and percent of normal are computed against the 1991–2024
  climatology — ERA5 for the observed and combined primera, and the SEAS5 July-issuance reforecast
  (1991–2024 hindcast) for the postrera forecast. Rainfall return periods are empirical Weibull ranks over the full available
  ERA5/SEAS5 series. ASI percentiles and return periods are taken against each unit's own 1984–2025
  FAO ASI record.
- **Confidence**: the primera rests on observed conditions and is high confidence; the postrera
  rests on a low-skill SEAS5 forecast, which supports anticipatory action but not a firm prediction.
- **Off-calendar**: the drought window falls outside the crop's rain-critical period (ASAP PSP).
  These units are better served by a vulnerability and food-security assessment than a
  seasonal-drought one.
- **Data and reproducibility**: the maps and table are built from derived files hosted on the team
  blob (`projects` container, `ds-aa-lac-dry-corridor/book/static/`), read at render time. The scripts
  that produce them — the database, Earth Engine, and raster pulls and the downstream transforms — are
  tracked in `data-raw/mission_support_2026/`.
- The vulnerability and food-security layer for the off-calendar areas has not yet been built.
