---
title: email
output: blastula::blastula_email
---

<style type = "text/css">

h1, h2, h3 {
  font-family: Arvo, "Helvetica Neue", Helvetica, Arial, sans-serif;
  font-weight: normal
}

p {
  font-family: "Source Sans Pro", "Helvetica Neue", Helvetica, Arial, sans-serif
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fontfamily = "Sans Source Pro",
  results="asis",
  fig.align='center',
  fig.cap = "" 
)
```

```{r}
# gghdx()
# source("src/email/email_utils.R")
```

```{r, class.output = "banner"}
# box::use(cumulus)
# box::use(blastula[...])
tfp_banner <- cumulus$blob_read(
  container = "projects",name = "ds-aa-lac-dry-corridor/framework_update_2025/email_logos/centre_banner.png" , return_path = T,progress_show = F
)
# cumulus$blob_read()
# banner_drib <- drive_dribble %>%
#   filter(name == "centre_banner.png")
# drive_download(banner_drib, path = banner_path <- tempfile(fileext = ".png"))

add_image(
  file = tfp_banner,
  alt = "Centre for Humanitarian Data banner",
  align = "left",
  width = "25%"
)
```


# `r email_txt$title`

## `r email_txt$subtitle`

### `r email_txt$date_header`  `r email_txt$status`

```{r}
# m_choro

eu$add_ggplot_custom(
  plot_object = m_choro,
  alt = "Alert Map",
  height = 3.75,
  width = 5.5,
  html_width = 650
)
```


## `r email_txt$description_title`

`r email_txt$description_content`

```{r PlotDischarge}
eu$add_ggplot_custom(
  plot_object = p_rainfall,
  alt = "Alert plot",
  height = 3.75,
  width = 5.5,
  html_width = 650
)

```

```{r tblThresholds}
gt_threshold_table 

```


```{r tblAOI}
gt_aoi
```


**Forecast data source:** `r email_txt$data_source`

Data accessed `r trimws(format(lubridate$as_date(Sys.Date()), "%B %Y"))`. 

`r email_txt$ref_github`

----

## Contact

`r email_txt$contact_info`

----

```{r FooterLogo}


# logo_drib <- drive_dribble %>%
#   filter(name == "ocha_logo_wide.png")
# 
# drive_download(logo_drib, path = logo_path <- tempfile(fileext = ".png"))
tfp_logo <- cumulus$blob_read(
  container = "projects",
  name = "ds-aa-lac-dry-corridor/framework_update_2025/email_logos/ocha_logo_wide.png" , 
  return_path = T, progress_show = FALSE
)
add_image(
  file = tfp_logo,
  align = "center",
  width = "25%"
)
```

<p>
<center>

<b> OCHA Centre For Humanitarian Data </b>

Fluwelen Burgwal 58 | 2511 CJ The Hague | The Netherlands

</center>
</p>
