---
title: "email"
output: blastula::blastula_email
---

<style type = "text/css">

h1, h2, h3 {
  font-family: Arvo, "Helvetica Neue", Helvetica, Arial, sans-serif;
  font-weight: normal
}

p {
  font-family: "Source Sans Pro", "Helvetica Neue", Helvetica, Arial, sans-serif
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fontfamily = "Sans Source Pro",
  results="asis",
  fig.align='center',
  fig.cap = "" 
)
```

```{r}
# gghdx()
# source("src/email/email_utils.R")
```

```{r, class.output = "banner"}
banner_drib <- drive_dribble %>%
  filter(name == "centre_banner.png")
drive_download(banner_drib, path = banner_path <- tempfile(fileext = ".png"))

add_image(
  file = banner_path,
  alt = "Centre for Humanitarian Data banner",
  align = "left",
  width = "25%"
)
```


# `r email_txt$title`

## `r email_txt$subtitle`

### `r email_txt$date_header`  `r email_txt$status`

```{r}
# m_choro
add_ggplot_custom(
  plot_object = m_choro,
  alt = "Alert Map",
  height = 3.75,
  width = 5.5,
  html_width = 650
)
```


## `r email_txt$description_title`

`r email_txt$description_content`

```{r PlotDischarge}
add_ggplot_custom(
  plot_object = p_rainfall,
  alt = "Alert plot",
  height = 3.75,
  width = 5.5,
  html_width = 650
)

```

```{r tblThresholds}
gt_threshold_table %>% 
  gt::tab_options(
    table.font.size = 14,
    # table.width = px(500)
    table.width = pct(80)
  )
```

**Forecast data source:** `r email_txt$data_source`

Data was accessed on `r trimws(format(as_date(Sys.Date()), "%e %B %Y"))`. 

`r email_txt$ref_github`

----

## Contact

`r email_txt$contact_info`

----

```{r FooterLogo}


logo_drib <- drive_dribble %>%
  filter(name == "ocha_logo_wide.png")

drive_download(logo_drib, path = logo_path <- tempfile(fileext = ".png"))

add_image(
  file = logo_path,
  align = "center",
  width = "25%"
)
```

<p>
<center>

<b> OCHA Centre For Humanitarian Data </b>

Fluwelen Burgwal 58 | 2511 CJ The Hague | The Netherlands

</center>
</p>
