---
title: "Joint Activation Rates"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  include: true
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
  code-fold: true
editor: visual
project:
  execute-dir: project
---

# Joint Activation Rates

Code takes output from `_targets.R` pipeline to create nicely formatted tables with joint activation rates for sharing.

The only table being shared is the one under `ECMWF + INSUV Combined`, the rest are merely for the interest/convenience of analyst.

For final analysis of joint activation rates we must consider the final approved activation moments:

-   **Primera (MJJA):** March-April-May
-   **Postera (SON):** June - July- August

The full discussion on how the above dates were chosen is out of scope of this document, but can be found in the [technical note](insert%20link) and [trigger framework](insert%20link)

```{r}
library(targets)
library(tidyverse)
library(gt)
library(here)
tar_load(df_insuvimeh_seasonal_summarised_filtered,store = here("_targets"))
tar_load(df_mars_seasonal_summarised, store = here("_targets"))
tar_load(df_all_thresholds_rp4,store = here("_targets"))

ldf_thresholds <- df_all_thresholds_rp4 %>% 
  split(.$forecast_source)
  
df_ecmwf_classified <- df_mars_seasonal_summarised %>% 
  filter(
    # not monitorng in Feb for Primera,
    # not monitoring May or Sep for Postera
    !(month(pub_date)==2 & window=="primera"),
    !(month(pub_date)==5 & window=="postera"),
    !(month(pub_date)==9 & window=="postera")
  ) %>% 
  left_join(ldf_thresholds$`ECMWF MARS`) %>% 
  
  mutate(
    flag = mm<= q_val,
    yr_pub_date = floor_date(pub_date,"year")
  ) 

df_joint_by_season_ecmwf <- df_ecmwf_classified %>% 
  group_by(
    adm0_es,window, yr_pub_date
  ) %>% 
  summarise(
    flag= any(flag),
  ) %>% 
  summarise(
    mean_flag = mean(flag),
    rp = 1/mean_flag
    
  ) %>% 
  arrange(
    window
  )



 df_insuv_classified<- df_insuvimeh_seasonal_summarised_filtered %>% 
   # not monitoring Feb for primera,
   # not monitoring May for postera
   
  filter(
    !(month(pub_date)==2 & window=="primera"),
    !(month(pub_date)==5 & window=="postera"),
    # just adding here for clarity, but already Sep is already filtered
    # INSIVUMEH only gives pub_mo + 1, no  pub_mo = valid_mo forecast
    !(month(pub_date)==9 & window=="postera") 
    ) %>% 
  left_join(ldf_thresholds$INSIVUMEH) %>% 
  
  mutate(
    flag = mm<= q_val,
    yr_pub_date = floor_date(pub_date,"year")
  ) 
 
df_join_by_season_insuv <- df_insuv_classified %>% 
  group_by(
    adm0_es,window, yr_pub_date
  ) %>% 
  summarise(
    flag= any(flag),
  ) %>% 
  summarise(
    mean_flag = mean(flag),
    rp = 1/mean_flag
    
  ) %>% 
  arrange(
    window
  )

```

## ECMWF Only

Joint activation across all lead times per season

```{r}
df_joint_by_season_ecmwf %>% 
  ungroup() %>% 
  gt()
```

Joint activation across all leadtimes and seasons

```{r}
df_ecmwf_either_window <- df_ecmwf_classified %>% 
  group_by(
    adm0_es, yr_pub_date
  ) %>% 
  summarise(
    flag= any(flag),
  ) %>% 
  summarise(
    mean_flag = mean(flag),
    rp = 1/mean_flag
    
  ) 
df_ecmwf_either_window %>% 
  gt()
```

## INSUV only

Joint activation across all lead times per season for just INSIVUMEH data

```{r}
df_join_by_season_insuv %>% 
  ungroup() %>% 
  gt()
```

joint activation if we were to look at either window with INSUV Data only

```{r}
df_insuv_either_window <- df_insuv_classified %>% 
  group_by(
    adm0_es, yr_pub_date
  ) %>% 
  summarise(
    flag= any(flag),
  ) %>% 
  summarise(
    mean_flag = mean(flag),
    rp = 1/mean_flag
    
  ) 
df_insuv_either_window %>% 
  gt()
```

## ECMWF + INSUV Combined

This are the joint activation rates that are the most relevant to the current framework which combines ECMWF & INSUV data as shown in the tables in `analysis/trigger_thresholds.qmd`

```{r}
# take classified data from INSUV and remove relevant Guatemala data that we will monitor with INSUVI
df_ecmwf_insuv_classified <- df_ecmwf_classified %>% 
  # We use INSUV for Guatemala except when LT = 0 as there 
  # is no LT = 0 in INSUV. This is just May in our framework
  filter(
    !(adm0_es=="Guatemala" & lt %in% 1:3)
  ) %>% 
  # Join insuv classified data
  bind_rows(
    df_insuv_classified
  ) 


#
df_ecmwf_insuv_joint_by_season <- df_ecmwf_insuv_classified %>% 
    group_by(
    adm0_es,window, yr_pub_date
  ) %>% 
  summarise(
    flag= any(flag),
  ) %>% 
  summarise(
    mean_flag = mean(flag),
    rp = 1/mean_flag
    
  ) %>% 
  arrange(
    window
  )
  # just for checking!
    # filter(adm0_es=="Guatemala", window=="primera") %>% 


df_ecmwf_insuv_joint_either <- df_ecmwf_insuv_classified %>% 
  arrange(window,pub_date) %>% 
    group_by(
    adm0_es, yr_pub_date
  ) %>% 
  summarise(
    flag= any(flag),
  ) %>% 
  summarise(
    mean_flag = mean(flag),
    rp = 1/mean_flag
    
  ) 
  
```

```{r}
# for nice formatting we can put by season and combined together into one table like this:

df_ecmwf_insuv_joint_by_season_wide <- df_ecmwf_insuv_joint_by_season %>% 
  mutate(
    # fix spelling for table publication
    window = str_to_title(
      case_when(
        window=="postera"~"Postrera",
        .default= window
      )
    )
  ) %>% 
  pivot_wider(
    names_from= adm0_es,
    values_from = rp:mean_flag
  )

df_ecmwf_insuv_joint_either_wide <- df_ecmwf_insuv_joint_either %>% 
    pivot_wider(
    names_from= adm0_es,
    values_from = rp:mean_flag
  ) %>% 
  mutate(
    window = "Combined",.before = everything()
  )

df_joint_activations_combined_wide <- bind_rows(
  df_ecmwf_insuv_joint_by_season_wide,
  df_ecmwf_insuv_joint_either_wide
  
)

df_joint_activations_combined_wide  %>% 
  mutate(
    window = fct_relevel(window, "Primera","Postrera","Combined")
  ) %>% 
  arrange(window) %>% 
  gt() %>% 
  fmt_percent(columns = starts_with("mean_"),decimals = 0) %>% 
  fmt_number(columns = starts_with("rp_"),decimals = 1) %>% 
   cols_merge(
    columns = c(`rp_El Salvador`, `mean_flag_El Salvador`),
    pattern = "{1} ({2})"
  ) %>% 
   cols_merge(
    columns = c(`rp_Nicaragua`, `mean_flag_Nicaragua`),
    pattern = "{1} ({2})"
  ) %>% 
   cols_merge(
    columns = c(`rp_Honduras`, `mean_flag_Honduras`),
    pattern = "{1} ({2})"
  ) %>% 
   cols_merge(
    columns = c(`rp_Guatemala`, `mean_flag_Guatemala`),
    pattern = "{1} ({2})"
  ) %>% 
  cols_label(
    "rp_El Salvador" = "El Salvador",
    "rp_Guatemala" = "Guatemala",
    "rp_Nicaragua" = "Nicaragua",
    "rp_Honduras" = "Honduras",
  ) %>% 
    tab_footnote(
      footnote = html("By continuously monitoring each publication date/leadtime with the same threshold (1 in 4 year return period level) the chance of an activation occuring across any leadtime increases. Furthermore, when more than 1 window is monitored in this manner, the chance of activating across either window also increases. Above we present this joint activation rates/RP levels using historical analysis of forecast data from ECMWF and INSIVUMEH data.")) %>% 
  tab_header(title = "Joint  Return Periods/Activation Rates",
             subtitle = "Across all monitored leadtimes and seasons"
             )
```
