---
title: "Central American Dry Corridor -Trigger Simulation"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 3
        code_folding: hide
date: "2023-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE, 
  eval = TRUE,
  results = "asis",
  out.width= "100%"
)
```

# Intro

In this document we simulate the trigger using historical IRI and historical VHI data. The purpose of this document is not performance testing/historical validation of thresholds. Rather, we simply visualize all historical VHI (FAO) and Forecast (IRI) data with respect to provided/proposed thresholds. If/when more optimal thresholds are identified we can update or append new visualizations using/modifying the code below.

```{r libsData,include=F}
library(targets)
library(tidyverse)
library(gghdx)
library(gt)
gghdx()
tar_load(df_cropland_lte_vhi_threshold)
tar_load(df_iri_adm0)

thresh_vhi <- 0.35
thresh_area <-  0.25
thresh_iri_pbavg <- 0.425
```


Initial triggers:

1. Average probability of below average rainfall is ≥ **`r round(thresh_iri_pbavg*100,1)`%***
2. ≥ **`r round(thresh_area*100,0)` %** of cropland with VHI ≤ **`r thresh_vhi`** - only monitored for Primera (phase 2) and Postera (phase 3) - observational


```{r}

# take max probability (bavg) across leadtimes
df_iri_adm0_max <- df_iri_adm0 %>% 
  group_by(
    across(starts_with("adm")), 
    predict_start_mo, seas
  ) %>% 
  summarise(
    across(.cols= c("mean","median"), ~max(.x/100)),.groups="drop"
  ) 

# filter to just relevant seasons
df_iri_adm0_max_relevant_season <- df_iri_adm0_max %>% 
  ungroup() %>% 
  filter(
    str_detect(seas,"MJJ|JJA|SON")
  ) 
```


# IRI

```{r}
# prep IRI data - create boolean flag indicating if threshold was crossed
df_iri_adm0_flag <- df_iri_adm0_max_relevant_season %>% 
  group_by(across(starts_with("adm")),seas) %>% 
  mutate(
    flag_thresh = mean>=thresh_iri_pbavg
  ) %>% 
  ungroup() %>% 
  mutate(
    # for tile plots it will be better have equal interval breaks on x-axis rather than dates
    # especially when we overlay the vhi data as a transparent tile
    date_cat = dplyr::dense_rank(predict_start_mo)
  )
```


The table below shows the mean probability of a below average rainfall for each "monitored" season that IRI data is available historically. The cell values are colored by the amount the exceed the current proposed threshold of  **`r round(thresh_iri_pbavg*100,1)`%***

```{r}
df_iri_adm0_flag %>% 
  select(adm0_es,Season=seas,mean) %>% 
  pivot_wider(
    names_from = adm0_es,
    values_from=mean
  ) %>% 
  # i thought `gt()` would be cool to use - but the solution i came up 
  # w/ below is a bit tedious
  gt() %>% 
  fmt_number(decimals=2) %>% 
  data_color(
    columns= `Guatemala`,
    rows = `Guatemala`>=thresh_iri_pbavg,
    palette= c(hdx_hex("tomato-light"), hdx_hex("tomato-dark"))
  ) %>% 
  data_color(
    columns= `El Salvador`,
    rows =`El Salvador`>=thresh_iri_pbavg,
    palette= c(hdx_hex("tomato-light"), hdx_hex("tomato-dark"))
  ) %>% 
  data_color(
    columns= `Nicaragua`,
    rows =`Nicaragua`>=thresh_iri_pbavg,
    palette= c(hdx_hex("tomato-light"), hdx_hex("tomato-dark"))
  ) %>% 
  data_color(
    columns= `Honduras`,
    rows =`Honduras`>=thresh_iri_pbavg,
    palette= c(hdx_hex("tomato-light"), hdx_hex("tomato-dark"))
  )

```

Below is another representation of the same data, just showing the timeline and displaying where and when the IRI threshold would have been exceeded.

```{r}
fill_cols <- c("white",hdx_hex("mint-hdx"))

df_iri_adm0_flag %>% 
  mutate(
    date_cat = dplyr::dense_rank(predict_start_mo)
  ) %>% 
  ggplot(
    aes(x= date_cat,y=adm0_es,fill=mean)
  )+
  geom_tile(color="lightgrey")+
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom", 
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
               guide = "bins",
               breaks = c(0,thresh_iri_pbavg,1), 
    
      )+
  scale_x_continuous(
    breaks= unique(df_iri_adm0_flag$date_cat),
    labels= unique(df_iri_adm0_flag$seas)
  )+
    labs(
    title = "Dry Corridor Central America",
    subtitle= paste0("Green cells indicate where the average probability of a below average seasonal rainfall was ≥ ",thresh_iri_pbavg)
  )+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    legend.position="none",

  )
```

# VHI

```{r}
# VHI is observational and involves monitoring the previous month of data

df_vhi_season <- df_cropland_lte_vhi_threshold %>% 
  mutate(
    date= ym(date),
    monitoring_window = case_when(
      # Primera 2 is JJA- therefore monitoring MJJ
      month(date) %in% c(5,6,7) ~ "Primera (phase 2) Monitoring",
      
      # Primera 3 is SON - therefore monitoring ASO
      month(date) %in% c(8,9,10) ~ "Postera (phase 3) Monitoring",
      .default=NA
    )
  ) %>% 
  filter(!is.na(monitoring_window)) %>% 
  mutate(
    monitoring_window = fct_relevel(monitoring_window,
                                    "Primera (phase 2) Monitoring",
                                    "Postera (phase 3) Monitoring")
  )


 df_vhi_seas_flagged <- df_vhi_season %>% 
   filter(
     round(threshold,2) ==thresh_vhi
   ) %>% 
   mutate(
     vhi_area_flagged = pct_crop_pixels_lte_thresh>= thresh_area
   ) %>% 
   filter(
     vhi_area_flagged
     ) %>% 
   group_by(across(starts_with("adm")), yr= year(date),monitoring_window) %>% 
   filter(date==min(date)) %>% 
   ungroup()
```
 
 
Below we visualize a timeline of all historical VHI data available by the 2 monitoring windows where VHI is a proposed trigger indicator. The y- axis is the % of cropland (per country) with a VHI≤ `r thresh_vhi`.  Years where the % cropland area (`r round(thresh_area*100,0)`) breached are colored in blue.
```{r}
 p_line_vhi_historical <- df_vhi_season %>% 
   filter(
     round(threshold,2) ==thresh_vhi
   ) %>% 
   mutate(
     pt_color = pct_crop_pixels_lte_thresh>= thresh_area
   ) %>% 
   ggplot(aes(
     x= date,
     y=pct_crop_pixels_lte_thresh
     
   ))+
    geom_hline(
    yintercept = thresh_area,
    color= hdx_hex("sapphire-hdx"),
    alpha= 0.3
  )+
   geom_vline(
     data= df_vhi_seas_flagged,
     aes(xintercept = date),linetype= "dashed",
     color = hdx_hex("tomato-hdx"),
     lwd=0.4,
     alpha=0.5
   )+
   geom_point(
     aes(
       color=pt_color
     ), size=0.8
   )+
   geom_line(color="grey")+
   scale_x_date(
     date_breaks = "1 year",
     date_labels = "%y",
     expand = c(0.01,0.01)
   )+
   geom_text(
     data=  df_vhi_seas_flagged,
     aes(
       x= date,
       y=0.6,
       label = format(date, "%y"),
       
     ),angle=90,
     color= hdx_hex("tomato-dark"),
     size=2
     
   )+
   labs(
     y= "% Cropland",
     title = paste0("Dry Corridor - Central America: % Cropland with VHI ≤ ",thresh_vhi),
     subtitle = "Red lines indicate seasons where trigger would be activated (threshold = 25%)"
   )+
   scale_y_continuous(
     labels = scales::label_percent()
   )+
   
   # facet_grid(rows=vars(monitoring_window), cols = vars(adm0_es))
   facet_grid(cols=vars(monitoring_window), rows = vars(adm0_es))+
   theme(
     panel.border = element_rect(color="grey",fill = NA),
     legend.position = "none",
     axis.title.x = element_blank(),
     axis.text.x = element_text(angle=90,size=7),
     axis.text.y = element_text(size=7)
   )
 

 p_line_vhi_historical

```


```{r}

df_vhi_thresh<- df_vhi_season %>% 
  # glimpse() %>% 
   ungroup() %>% 
  mutate(
    season_gen = case_when(
      monitoring_window== "Primera (phase 2) Monitoring"~"JJA",
      monitoring_window== "Postera (phase 3) Monitoring"~"SON",
      .default = NA
      ),
    mo_start = ifelse(season_gen== "JJA","06","09"),
    predict_start_mo = ym(paste0(year(date),"-",mo_start)),
    # year = year(date),
    # yr = format(year,"")
    seas = paste0(season_gen, format(date,"%Y"))
  ) %>% 
  filter(
    !is.na(season_gen),
    round(threshold,2)==thresh_vhi
    ) %>% 
  group_by(adm0_es,yr=year(predict_start_mo),monitoring_window) %>% 
  summarise(
    pct_max= max(pct_crop_pixels_lte_thresh,na.rm = T),
    .groups="drop"
  ) 
```

The same data is represented in a simplified heat plot below. The heat plot cell values are colored by the % cropland breaching the propsed VHI threshold and years which breach the % cropland threshold are outlined in red.
```{r}
df_vhi_thresh %>% 
  ggplot(
    aes(x= yr,
        y=adm0_es,
        fill=pct_max)
  )+
  geom_tile(color="lightgrey")+
  geom_tile(
    data= df_vhi_thresh %>% 
      # group_by(adm0_es, yr, monitoring_window) %>% 
      filter(pct_max>=thresh_area),
    color=hdx_hex("tomato-hdx"),
    lwd=0.5,
    fill=NA
  )+
  scale_fill_continuous(name= "% cropland",
                        labels=scales::label_percent()
                        )+
  facet_wrap(~monitoring_window,nrow=2)+
  scale_x_continuous(
    breaks= unique(df_vhi_thresh$yr)
  )+
    labs(
    title = "Dry Corridor Central America",
    subtitle= "% Croplandwith VHI ≤ 0.35 - Years with ≥ 25 % highlighted in red"
  )+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    legend.text = element_text(angle=90)
  )

```

# IRI + VHI Combined

As the framework proposes using both triggers (IRI & VHI) - let's see how they line up for all the years that we have both historically.

```{r}
fill_cols <- c("white",hdx_hex("mint-hdx"))
# let's align vhi  to the IRI seasons (df_iri_adm0_flag) for vizualization

df_vhi_seas_aligned <- df_vhi_seas_flagged %>% 
  ungroup() %>% 
  mutate(
    season_gen = case_when(
      monitoring_window== "Primera (phase 2) Monitoring"~"JJA",
      monitoring_window== "Postera (phase 3) Monitoring"~"SON",
      .default = NA
      ),
    mo_start = ifelse(season_gen== "JJA","06","09"),
    predict_start_mo = ym(paste0(year(date),"-",mo_start)),
    # year = year(date),
    # yr = format(year,"")
    seas = paste0(season_gen, format(date,"%y"))
  ) %>% 
  filter(
    year(date)>= 2017
  ) %>% 
  select(starts_with("adm"),predict_start_mo,seas,threshold,pct_crop_pixels_lte_thresh)

```


```{r}
df_iri_adm0_flag %>% 
  ggplot(
    aes(x= date_cat,y=adm0_es,fill=mean)
  )+
  geom_tile(
    color="lightgrey",lwd=0.1
    )+
  
  geom_tile(
    data= df_vhi_seas_aligned %>% 
      # this join bring in the `date_cat` (dens_rank(date)) values in
      left_join(
        df_iri_adm0_flag
      ),

    fill=NA,
    color=hdx_hex("tomato-hdx"),
    alpha=0.5, lwd=0.5
  )+
  
  binned_scale(
    aesthetics = "fill",
    scale_name = "custom", 
    palette = ggplot2:::binned_pal(scales::manual_pal(values = fill_cols)),
               guide = "bins",
               breaks = c(0,thresh_iri_pbavg,1), 
    
      )+
  labs(
    title = "Dry Corridor Central America",
    subtitle= paste0("Green cells indicate where probability of below average rainfall was ≥ ",
                     round(thresh_iri_pbavg*100,1),
                     " %\nRed outline indicates when ≥ ",
                     round(thresh_area*100,0),
                     " % of cropland had a VHI ≤ ",
                     thresh_vhi)
  )+
  scale_x_continuous(
    breaks= unique(df_iri_adm0_flag$date_cat),
    labels= unique(df_iri_adm0_flag$seas)
  )+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    legend.position="none"
  )

```

